\documentclass{article}
\usepackage{fontspec,multicol,xcolor,array,xspace,pgf,ifxetex}
\newfontfamily\meitei{NotoSansMeeteiMayek-Regular.ttf}
\newfontfamily\pan{code2000.ttf}
\newfontfamily\smallcps[Scale=0.8]{Arial Unicode MS}
\newfontfamily\brahmi{Noto Sans Brahmi}
\DeclareRobustCommand\unicodenumber[1]{{\smallcps #1\xspace}}
\def\textmeitei#1{{\meitei #1}}
\newfontfamily\aegean{Aegean.ttf}
\newfontfamily\arial{Arial Unicode MS}
\makeatletter
%    \begin{macrocode}
\def\graybox{x}
\newfontfamily\unicodenumberfam{Arial Unicode MS}
\def\textU#1{{\unicodenumberfam #1}}
\makeatletter
\def\putunicode@label#1#2;{%
\def\reformat@unicode@string##1{%
   \textU{U+}%
  \let\z\empty%
  \expandafter\@tfor\expandafter\i\expandafter:\expandafter=#2;\do{%
  \if\i;%
    \textU{x}%
  \else%
    \textU{\z}%
  \fi%
  \edef\z{\i}%
 }%
}%
  \makebox[5em]{\reformat@unicode@string{#2}\hfill}%
}

% \end{macro}
% 1#  text file #2 font command
%    \begin{macrocode}
\def\putchar@cx#1{%
   \iffontchar\font\n
     \char\the\n$_{\pgfmathparse{Hex(\the\r@cx)}\textsf{\pgfmathresult}}$%
      %
   \else
     \graybox
   \fi
\let\active@prefix\oldactive@prefix
 }

\def\urow@cx#1{%
    \n=#1% 
    \r@cx=0%
    \expandafter\putunicode@label#1;%
    \loop%
        \ifnum\n<\numexpr#1+16\relax%
        \makebox[2.1em]{\expandafter\putchar@cx{#1}}%
        \advance\r@cx by1%  
        \ifnum\r@cx>16\r@cx=1\relax\else\fi
        \advance\n by1%
    \repeat
    \par
}

\def\typeseturows@cx#1{%
\@for\next:=#1\do{%
  \urow@cx\next\vskip3pt}%
}

\newcount\r@cx%
\newcount\n%
\newcommand\unicodetable[2]{%
\bgroup
  \par
  \leavevmode%
   \r@cx=0%
   {\hbox to 5em{\ignorespaces}}%
   \loop%
    \ifnum\r@cx<16\ignorespaces 
    \makebox[2.1em]{\pgfmathparse{Hex(\the\r@cx)}\pgfmathresult}%
    \advance\r@cx by\@ne%  
   \repeat
   \vskip3pt\par
   \@nameuse{#1}%
   \typeseturows@cx{#2}%
\egroup
}

%    \end{macrocode}

\def\putdescription#1:{%
  \parindent0pt 
  \begin{minipage}[t]{4cm}
  \bgroup\aegean
  \hangindent20pt
  #1\par
  \egroup
  \end{minipage} 
}
% \end{macro}

\long\def\parsefields #1:#2\@@{%
    \ifx\par#1
    \else 
        {\small\aegean U+#1}%
         %\iffontchar\font"#1 %
          \makebox[2.1em]{\color{blue}\symbol{"#1}}% 
          \expandafter\putdescription#2\vskip3pt
        %\else
          %{\aegean \makebox[2.1em]{} Unallocated\par}%
        %\fi
    \fi  
  }%


\newread\unicodefile

% begin{macro}{\printunicodeblock}\marg{}\oarg{} The macro
% prints a unicode table from a file of definitions. This is
% printed in a two column environment by default. 
% #1 number of columns (optional) 
% #1 filename and path
% #2 font command
\DeclareDocumentCommand{\printunicodeblock}{O{2} m m }{%
  \bgroup
  \leavevmode\parindent0pt\par
  \begin{multicols}{#1}%
  #3
  \openin\unicodefile=#2
  \loop\unless\ifeof\unicodefile
    \read\unicodefile to\fileline %
    \expandafter\parsefields \fileline:\@@ 
  \repeat
  \end{multicols}%
  \egroup
  \closein\unicodefile}


\begin{document}
\arial


Meithei (Meitei) /ˈmeɪteɪ/,[4] also known as Manipuri /mænɨˈpʊəri/ ({\pan মৈতৈলোন্} \textmeitei{ꯃꯧꯇꯧꯂꯣꯟ} Meitei-lon or {\pan মৈতৈলোল্} \textmeitei{ꯃꯧꯇꯧꯂꯣꯜ} Meitei-lol), is the predominant language and lingua franca in the southeastern Himalayan state of Manipur, in northeastern India. It is the official language in government offices. Meithei is also spoken in the Indian states of Assam and Tripura, and in Bangladesh and Burma (now Myanmar).

Meithei is a Tibeto-Burman language whose exact classification remains unclear, though it shows lexical resemblances to Kuki and Tangkhul Naga.[5] The language is spoken by more than 1.5 million people.

Meithei has proven to be an integrating factor among all ethnic groups in Manipur who use it to communicate among themselves. It has been recognized (as Manipuri), by the Indian Union and has been included in the list of scheduled languages (included in the 8th schedule by the 71st amendment of the constitution in 1992). Meithei is taught as a subject up to the post-graduate level (Ph.D.) in universities of India, apart from being a medium of instruction up to the undergraduate level in Manipur.

\bgroup
\meitei
\begin{tabular}{>{\arial}l
                >{\arial}l
                >{\meitei}l
                >{\arial}l
                >{\arial}l
                >{\meitei}l
               }
1	&ama 	 &ꯑꯃ	       &11	&taramathoi	&\\
2	&ani	   &ꯑꯅꯤ	&12	 &taranithoi	&{\arial ky} \\
3	&ahum	&ꯑꯍꯨꯝ	   &13	 &tarahumdoi	&ꯇꯔꯥꯍꯨꯝꯗꯣꯢ\\
4	&mari	&ꯃꯔꯤ	   &14  &	taramari	&ꯇꯔꯥꯃꯔꯤ\\
5	&manga	 &ꯃꯉꯥ	   &15	 &taramanga	&ꯇꯔꯥꯃꯉꯥ\\
6	&taruk	 &ꯇꯔꯨꯛ	   &16	 &tarataruk	&ꯇꯔꯥꯇꯔꯨꯛ\\
7	&taret	 &ꯇꯔꯦꯠ	   &17	 &tarataret	&ꯇꯔꯥꯇꯔꯦꯠ\\
8	&nipan &ꯅꯤꯄꯥꯟ	&18	 &taranipan	&ꯇꯔꯥꯅꯤꯄꯥꯟ\\
9	&mapan	 &ꯃꯥꯄꯟ	   &19	 &taramapan	&ꯇꯔꯥꯃꯥꯄꯟ\\
10	&tara	 &ꯇꯔꯥ	   &20	 &kun	&ꯀꯨꯟ\\
\end{tabular}
\egroup

Meitei Mayek script was added to the Unicode Standard in October, 2009 with the release of version 5.2.

The Unicode block for Meitei Mayek, called Meetei Mayek, is \unicodenumber{U+ABC0–U+ABFF}.

Characters for historical orthographies are part of the Meetei Mayek Extensions block at \unicodenumber{U+AAE0–U+AAFF}.

%\begin{scriptexample}[]{Meitei}
%\unicodetable{meitei}{"ABC0,"ABCD0,"ABE0,"ABF0}
%\end{scriptexample}
%
%\begin{scriptexample}[]{Meitei}
%\unicodetable{meitei}{"AAE0,"AAF0}
%%\captionof{table}{Meetei Mayek Extensions}
%\end{scriptexample}

%\begin{multicols}{2}
\printunicodeblock{./languages/meetei-mayek.txt}{\meitei}
%\end{multicols}

\section{Brahmi}
\label{s:brahmi}
Brāhmī is the modern name given to one of the oldest writing systems used in the Indian subcontinent and in Central Asia during the final centuries BCE and the early centuries CE. Like its contemporary, Kharoṣṭhī, which was used in what is now Afghanistan and Western Pakistan, Brahmi (native to north and central India) was an \emph{abugida}.

The best-known Brahmi inscriptions are the rock-cut edicts of Ashoka in north-central India, dated to 250–232 BCE. The script was deciphered in 1837 by James Prinsep, an archaeologist, philologist, and official of the East India Company.[1] The origin of the script is still much debated, with current Western academic opinion generally agreeing (with some exceptions) that Brahmi was derived from or at least influenced by one or more contemporary Semitic scripts, but a current of opinion in India favors the idea that it is connected to the much older and as-yet undeciphered Indus script



Brahmi is a Unicode block containing characters written in India from the 3rd century BCE through the first millennium CE. It is the predecessor to all modern Indic scripts.

%\begin{scriptexample}[]{Brahmi}
\unicodetable{brahmi}{"11000,"11010,"11020,"11030,"11040,"11050,"11060,"11070}
%\end{scriptexample}


\printunicodeblock[2]{./languages/brahmi.txt}{\brahmi}

\input{./languages/sumero-akkadian}

\newfontfamily\samaritan[RawFeature=samr]{NotoSansSamaritan-Regular.ttf}


\printunicodeblock{./languages/samaritan.txt}{\samaritan}

























\end{document}