\documentclass[twoside]{book}
\usepackage{lipsum}

\begin{document}
We refer to the Engineer’s letter Ref:.........................., dated .................  copied to us via your letter Ref.... dated ....
and comment as follows:

\begin{enumerate}
\item The letter refers to co-ordination issues for works which we have not as yet been instructed to
execute but merely instructed to co-ordinate as per Engineer’s letter \texttt{DU1211/DU/L/20834/14} dated 29 September 2014 by reserving space for the busbar proposed to serve the Theatre. This we have done and we attach copies of these drawings for the record. The same drawings were issued at our meeting with the Engineer’s Electrical representatives as described in the Engineer’s letter.

\item We have written to the Engineer highlighting that the \textit{latest} drawings issued to us at the beginning of January via reference 28536 dated 30 Dec 2014  are in conflict with the original instruction and have requested that detailed designs be provided to us for further co-ordination.

\item It appears that the Engineer’s Electrical Representative is particularly sensitive to providing such information, and unfairly accuses the Contractor of writing unprofessional letters of questionable character and that all costs etc of unknown contnent would be paid by the Contractor. We note that the Project would have been better served by a) providing designs synchronously with the works on site, rather than in an adhoc and piecemeal manner and b) that all primary design co-ordination---which falls under the scope of works of the Designers---was concluded at the preliminary design stages as is the norm in the Industry rather than carried out at such a late stage.
Levelling accusations towards the Contractor of “unprofessionalism” cannot and will not resolve design issues. We understand that in a multiculture environment, language barriers can sometimes cause miscommunication but we find that the tone and character of the letter contrary to the spirit of team work that is necessary for the sucessful completion of the Project.
\end{enumerate}

We have assisted the Designers by modifying our drawings and installed services to comply with the instructions forwarded to us in September 2014. Unfortunately we are unable to provide any further assistance in this respect until such time as a) the information we have requested in our letter is provided.



\makeatletter
\let\ol@outputpage\@outputpage
\def\@outputpage{%
    \ifodd\c@page\else\shipout\vbox{}\advance\c@page\@ne\fi
\ol@outputpage}
\makeatother

\fbox{\parbox{3in}{1\vskip .7\textheight}}

\fbox{\parbox{3in}{2\vskip .7\textheight}}

\fbox{\parbox{3in}{3\vskip .7\textheight}}

and back to normal:

%\makeatletter
%\let\@outputpage\ol@outputpage
%\makeatother
%
%\fbox{\parbox{3in}{4\vskip .7\textheight}}
%
%\fbox{\parbox{3in}{5\vskip .7\textheight}}
%
%\fbox{\parbox{3in}{6\vskip .7\textheight}}


\end{document}

http://tex.stackexchange.com/questions/57542/switch-to-one-sided-mode-in-the-appendix-of-a-double-sided-document/57617#57617