\chapter{Delay in Finalization of Air Handling Unit, Fans and Fan Coil Units}
\label{ch:ahus}

As per the Approved Baseline Program we planned to place the HVAC equipment orders on 20 March 2014. Changes to all equipment continued well into 2015. \idxahu{general delays}

\begin{enumerate}
\item On 17 September 2013, we were issued via \DT{0320} updated Mechanical Drawings. \CAR{0002}. This was the first major drawing issue, kick-starting the Engineering and Procurement process.

\item On 10 December 2013 we received updated Mechanical Drawings which revised many of the previously issued drawings and equipment schedules. (See \CAR{0001} and \CAR{0002})

\item On 11 February 2014, via  \DT{0813} we were issued updated Mechanical Drawings for Basements which led to  abortive Engineering works and precluded the finalization of any detailed calculations and the procurement of equipment. (\CAR{0026})\basement{HVAC}

\item On 27 February 2014, we received via \DT{0878} updated Fire Fighting and HVAC drawings (28 Nos) for Basements which superceded some of the previously issued drawings (See \CAR{0031}.)\basement{HVAC}

\item On 8 March 2014 we received instructions via \DT{0898} and updated Mechanical drawings for all three hotels \CAR{0031}. 

\item On 16 March 2014 we received instructions via \DT{920} and updated Mechanical drawings for all three hotels.
\idxwestin{AHU}\idxwestin{fans}\idxwestin{fcu}.

\item On 10 April 2014 we received instructions via \DT{1027} ``Revised Mechanical Drawings'' with again more changes and revisions to the design.\idxahu{10 April 2014}

\item On 28 May 2014 we received more revisions via \DT{1194} ``Revised Mechanical Drawings for St Regis Hotel''.

\item On 13 July 2014 we received major changes to equipment schedules via email. 

\item On 4 August 2014 we issued letter \letter{1489} which was affecting the ordering of fans for the St. Regis Hotel and Basement areas.\basement{fans}

\item On 13 August 2014 \idxahus{13 Aug 2014 revisions} we received updated Mechanical Drawings again with equipment revisions. We reserved our rights via \letter {1504}. (See \CAR{0077})

\item On 1st September  2014 we received advance copies of further changes to the HVAC works. These were issued to us officially on 10 September 2014 and our confirmation followed by letter \letter{1549} and \CAR{0085}. 

\begin{enumerate}
  \item Additional AHUs were introduced.
  \item Additional fcus were introduced.
  \item Additional fans were introduced.
  \item Additional CAVs were introduced.
  \item Additional ductwork and revisions to duct sizes
  \item Additional piping.
  \item The revisions impacted on related Electrical works, Fire Alarm, Fire Detection and BMS systems. 
\end{enumerate}

The above resulted to numerous abortive works on site and voiding previous Engineering works.

\item On 25 September 2014 we reserved our rights to further revisions that we received via RFI replies. RFI/MEP/M307. \CAR{0085}\idxahus{25 Sept revisions}

\item On 2 October 2014, we confirmed additional works due to revised Mechanical Equipment Schedules. These were received via RFI/MEP/M299 for the St Regis Hotel and Basement levels \CAR{0085}.\idxbasement{equipment revisions}

\item Engineer requested us to upgrade one AHU for W Hotel and this was confirmed via letter \letter{2014}.\CAR{0085}.
\end{enumerate}

The above changes seriously impacted on the ordering, but most importantly on the finalization of critical equipment and ductwork for the whole development. In order to minimize the impact we employed a full time Senior Engineer just to handle these continuous changes and to manage the calculations for static pressures.











