
\chapter{Delays in Finalizing Requirements for the Busbar System}

As per the approved Baseline Program\footnote{Issued 4 Jan 14 and approved 9 Jan 14, as per HLG letter Ref: HLG/626/2.5/SO/nd/1862/14},  we were planning to order the Busbar on 27 February 2014. The \JV was unable to finalize the Bus Bar Material Submittal due to the numerous revisions issued and the lack of Dewa approved drawings.

\begin{enumerate}
\item On 23 December 2013 we received HLG transmittal ref: no. HLG-626-DT-HLS-0628 dated 23 December 2013 ``Fire Resistance Bus Bar Specification'', instructing us to change some of the busbars to fire rated busbars.
\idxbusbar{change in specification}\idxbusbar{fire rated}
\label{fireratedbusbar}

\item HLG transmittal Ref: No.: HLG-626-DT-HLS-0797 dated 10 February 2014 titled ``Electrical Updated Coordinated Drawings for Basements". (\CAR{0004}).

\item HLG Transmittal Ref. No.: HLG-626-DT-HLS-0873 dated 25 February 2014 ``Revised Electrical Drawings''.

\item On 18 and 20 March 2014 via \DT{0930\&939} we were issued updated drawings for three Hotel (\CAR{0036}).  

\item \DT {10127} dated 10 April 2014 ``Revised Electrical Drawings''. 

\item On 16 June 2014 we were instructed to stop any works on Westin and W Hotel Bus bars due to ``significant comments on Dewa LV approval''. This instruction was received via \KA letter ref. no. DU1211/DU/L/13086. At the time we had completed LV Schematic drawings and had the bus bar isometrics completed. 

\item The approved Dewa LV drawings were received on 15 July 2014 as per \DT {1397} dated 15 July 2014. One month later than the instruction to hold the orders and the works.

\item On 20 August 2014, we requested that we be issued officially the St Regis Hotel revised drawings that we were being send piecemeal by email. \letter{1515}.

\item On 1 September 2014 we had responded to HLG letter ref. no. HLG/626/1.12.AMM/es/4109/14 expressing concerns as to delays in receiving receiving workable design drawings (\letter{1533} (\CAR{0083}).

\item We received revised Electrical Drawings for Westin via \DT {1671} dated 8 September 2014.

\item On 17 September 2014 we received Electrical Drawings for W Hotel via \DT{1728}. 

\item On 27 September we served notice for additional costs and time due to revised electrical drawings related to Mechanical Equipment. The additional loads were received as response to RFI/MEP/E295 for all three Hotels (\CAR{0083}, \CAR{0086}\& \CAR{0087}).

\item On 2 October 2014 we served notices for additional works due to revised Electrical Drawings for Food and Beverage areas. (\CAR{0093}) {\CAR{0093}}

\item We served notice for drawings issued to us for W Hotel from 24 to 27 floor (\CAR{0112}).

\item On 19 February 2015 we issued estimated costs at the Engineer's request for the fire rated bus bars. On 25 March 2015 we were instructed to change all fire rated bus bars to normal bus bars. Accordingly the procurement of this particular bus bars took from 23 December 2013 until 19 February 2015 to be concluded and delayed the works.  (See \ref{fireratedbusbar} referred to when the first instruction was received. )


\end{enumerate}