\chapter{Dewa Approval}
\label{ch:dewa}

As per the approved Baseline Program, we were expected to receive Dewa approved drawings on the 28th November 2013. However, HLS-DSE JV received the LV approved drawings on 15th July 2014, as per HLG transmittal reference No. HLG-626-DT-HLS-1397 dated 15th July 2014. This delayed finalization of orders and progress on site.

 In particular:
 
 \begin{enumerate}
 \item Cables cannot be ordered until such time as approved single line diagrams are available. Once these become available  Shop drawings are prepared and main panels can also be finalized.
 \item MDBs and SMDBs can be finalized and ordered.
 \item Completion of Generator Rooms.
 \item Completion of Transformer and LV Rooms.
  \end{enumerate} 
  
\section{Action by the HLS DSE-JV}
  
Given the enormous task at hand and the instructions received to accelerate the works, we added an Electrical Engineering Manager to assist the Team with the task at hand. We also added additional CAD Operators.

\section{Design Deficiencies}

The Dewa drawings were out of step with the latest revisions of other drawings in terms of architectural, HVAC, Kitchen requirements and other equipment. They also underestimated both the main power required by 2.5MW, as well as the stand-by power required, leading to revisions to the Generator Plant. The Generator Plant is handled under delays of Electrical equipment.

Normally once drawings are submitted and approved by Dewa, the design can be considered complete, however, many areas remained incomplete.

\begin{enumerate}
\item On 20th August 2014, we requested by letter HLSDSEJV/HC/L/YL/1502 to be issued officially a number of revisions we received via email correspondence for the St Regis Hotel. 

\item On 21st August 2014 we confirmed receipt of revised Electrical Drawings from Ground to First Floor via letter ref. no. HLSDSEJV/HC/L/YL/1524. (\CAR{0076})\idxdewa{21 August revisions}

\item On 1 September 2014, we issued delay notice for revised electrical drawings, received by email for St. Regis via letter ref. no.: HLSDSEJV/HC/L/YL/1533 (CAR 83).\CAR{0083}

\item On 11 September, 2014 we confirmed via letter 

\item On 8 September 2014 we received further changes to Electrical Drawings for Westin via HLG Transmittal Ref. No. HLG-626-DT-HLS-1671 dated 8 September 2014.

\item As there was uncertainty over which drawings were to be used, HLG issued us a letter from the Engineer dated 11 September 2004, confirming the following:

      \begin{enumerate}
    	\item  St. Regis Hotel - Electrical Design drawings to be followed as per 1 September 2014 issue drawings (DU/L/18451/14).
    	\item Westin Hotel - Electrical Design drawings to be followed as per the 4 September issued drawings (DU/L/18896/14).
    	\item W Hotel - Electrical Design drawings will be issued after incorporating new Restaurant and ID drawings.
	  \end{enumerate}
\end{enumerate}

\section{30 December 2014 Dewa approved drawings issue}
\label{electrical}

On the 31 December 2014 final Dewa revised drawings were issued. This incorporated further revisons to electrical panels, additional SMDBs, changes to cable sizes, breaker sizes etc. Delay notice was served via letter Ref: YL/1796 date 19/1/2015. Changes affected all areas, including basements, St Regis, Westin and W Hotels. We wrote to the Engineer with suggestions to minimize the impact via letter Ref 25 January 2015 and recording the changes. For the W \& Westin Hotel we did the same via letter ref YL/1843 dated 3 February 2015 (\CAR {0136}).\CAR{0126}

The letters remained unanswered and we issued reminder letter related to these changes via letter ref YL/1907. We also confirmed that the works ere put on hold until such time as we had received confirmation from the Engineer.

Additional works as per letter \texttt{HLG/626/2.05/YE/es/7312/15} dated 6 April 2015. These changes relate to late approval of DEWA drawings. These changes affected all the hotels.

These revisions to the electrical design obstructed us from finalizing and ordering the Electrical Panels including MDBs, MCC, SMDB and electrical cables. The final impact of these changes is described below.

\section{St. Regis}
The following changes were instructed via the above letter and were based on drawing number |EM3300|.
\begin{description}
\item[SMDB-H1-1PLBPR] The works adds outgoing cables feeding |ADD-SS-01| and for |DBP-H1-1PLBPR1|  the cable size was changed from 4c:10mm2 XLPE to 4c:16 mm2 XLPE. The breaker size was changed to 60A MCCB.

\item[SMDB-H1-1TEFCWF] The instruction requests the changing of 15A breaker to 20A for eight CP-H1-1-TEWF/05 T.C.L.-1kW and one CP-H1-1TEWF/09 T.C.L.-1kW.

\item[SMDB-H1-2PL] The instruction requests the following changes:
   \begin{enumerate}
      \item DBP-H1-2PL MCCB 60A change to 80A and cable size 4c:16mm2 XLPE change to 4c:25mm2 XLPE.
      \item BPN-PN-16 and 18 30mA ELCB added.
   \end{enumerate}

\item[SMDB-H1-2PSPA] The instruction requests the following changes:
    \begin{enumerate}
      \item Male and female Jacuzzi bath MCCB 15A change to 20A TCL-3kW.
      \item Female steam room cable size changed (4c:10mm2 XLPE to 4c:16mm2 XLPE).
    \end{enumerate}


\item[SMDB-H1-6PL] The instruction requests the following changes:
   \begin{enumerate}
      \item Additional outgoing feeders for EC-01B, EC-02B, WET-PN-011, WET-PN-017 and WET-PN-020.
      \item 40ATP MCCB removed for FP-H1-1FL
   \end{enumerate}

\end{description}

The following changes were due to drawing No:EM3301

\begin{description}
\item [MDB-H1-B1R1] The instruction requests the following changes:
    \begin{enumerate}
       \item SMDB-H1-GR2 MCCB 200A change to 225A and cable size 4c:95mm2. XLPE change to 4c: 120mm2 XLPE (TCL 116.8kW).
       \item UPS MCCB 60A change to 80A.
    \end{enumerate}
\item[MDB-H1-GR2] The following changes were instructed:
    \begin{enumerate}
       \item Incomer MCCB 200A TP change to 225A TP.
       \item Additional outgoing for WPN-PA-012, WPN-PA-032.
    \end{enumerate}
\end{description}

The following changes were due to drawing No:EM3302

\begin{description}
\item[SMDB-H1-GLSTBR] The following changes were requested:
   \begin{enumerate}
      \item Additional outgping for St Regis, Special Event, St Regis BR.
      \item DBP-H1-GLSTBR MCCB 60A change to 80A.
   \end{enumerate}
\item[SMDB-H1-1PLMK] The following changes were requested:
      \begin{enumerate}
        \item Additional space.
        \item 60A TP MCCB removed.
      \end{enumerate}  
\item[SMDB-H1-2PGSC] The following changes were requested:
     \begin{enumerate}
        \item CAF-SS-01 cable and MCCB size changed from 4c:70mm2 XLPE and 150A TP to 4c:XLPE and 30A TP (TCL-6.5kW).
     \end{enumerate}
\end{description}

The following changes were detailed on drawing No:EM3303

\begin{description}
\item[MDB-H1-B1LR2] SMDB-H1-GL MCCB80A change to 100A.
\item[SMDB-H1-GL] DBP-H1-GLPFA MCCB 60A change 80A and cable size 4c:16mm2 XLPE change 4c:25mm2 XLPE.
\item[SMDB-H1-GLBP1] Additional outgoing for BOQ-KIT-016.
\end{description}

The following changes were detailed on drawing No:EM3304.

\begin{description}
\item[EMDB-H1-B1]  The following changes were requested:
   \begin{enumerate}
      \item ESMDP-H1-GR2 MCCB 80A change to 150A and cable size 4c:35mm2 XLPE change to 4c:70mm2 XLPE.
      \item ESMDB-H1-6PMS1 MCCB 400ATP change to 500ATP.
   \end{enumerate}
\item[EMDB-H1-6PMS1] The following changes were requested:
    \begin{enumerate}
       \item Incomer MCCB 400ATP change to 500ATP.
       \item Additional outgoing for EC-01A,B and future load.
       \item ESMDB-H1-RS cable size changed from 4c:70mm2 XLP (125A TP to 4c:95mm2 XLPE (TCL-55kW).
    \end{enumerate}
\item[ESMDB-H1-GL]
\item[ESMDB-H1-6PMS2]  The incomer to MCCB was changed from 700A TP to 800A TP.
\item[ESMDB-H1-6PL] An additional outgoing cable was requested for EC-02A. For LIFT-H1-SL05 and LIFT-H1-SL06 the cable size was requested to be changed to 4c:35mm2 MGT/XLPE.
\item[ESMDB-H1-2PL] CAF-SK-012, EC-01B MCCB and cable size changed from 30A SP 2c:16mm2 XLPE to 20ASP and 2c:4mm2 PVC (T.C.L.-2.6kW and 0.8kW).
\end{description}

\subsection{St Regis Basement Areas}
The following changes were detailed on drawing No:EM3200.\idxdewa{basements}\idxbasement{SMDB revisions}
\begin{description}
\item[SMDB-BP-1BS1] Additional outgoing circuits were requested for DB-LS-SR2, DB-LS-SR3.
\item[SMDB-BP-1BS3]  An additional outgoing circuit was instructed for DB-LS-SR5.
\item[SMDB-BP-1BS5] An additional outgoing circuit was requested for DB-LS-SR6.
\end{description}

The following changes were detailed on drawing No:EM3201.

\begin{description}
\item[EMDB-BP-1B3] ESMDB-BP-1BS7 MCCB 40A change to 80A and cable size 4c:10mm2 XLPE change 4c:16mm2 XLPE (TCL-17.8kW).\idxbasement{EMDB revisions}\idxbasement{ESMDB revisions}
\item[ESMDB-BP-1BS9] cable size 4c:35mm2 XLPE change to 4c:70mm2 XLPE (TCL-44.4kW).
\item[ESMDB-BP-1B3]  The following changes affected this panel:
     \begin{enumerate}
        \item ESMDB-BP-1BS7 MCCB 40A change to 80A and cable size 4c:10mm2 XLPE change to 4c:16mm2 XLPE    (TCL-17.8kW). 
        \item ESMDB-BP-1BS9 cable size 4c:35mm2 XLPE change to 4c:70mm2 XLPE (TCL-44.4kW). 
        \item ESMDB-BP1BS10 cable size 4c:240mm2 MGT change to 4c:300mm2 MGT(TCL-120kW).
     \end{enumerate}
\item[ESMDB-BP-1BS2]  3 Nos CP-BP-1BE/F1 cable size 4c:16mm2 MGT change to 4c:25mm2 MGT (TCL-17kW).
\item[ESMDB-BP-1BS3]  20ATP, pulse meter, 10mm2 MGT removed for SPCP-BP-1B12.
\item[ESMDB-BP-1BBPA] Incomer MCCB 80A TP change to 100A TP.
\item[ESMDB-BP-1BCOM1] Additional outgoing for COM-IC-001, COM-IC-002, COM-IC-003, COM-IC-006.
\item[USMDB-BP-1BS] UDB-BP-1BS4 and UDB-BP-1BS5 cable size 4c:10mm2 XLPE change to 4c:16mm2 XLPE (TCL-8.8kW and TCL-7.6kw).
\item[ESMDB-BP-1BS1] Incomer MCCB 200A TP change to 250A TP.
\item[ESMDB-BP-1B] ESMDB-BP-1BBPA MCCB 80A change to 100A and cable size 4c:50mm2 XLPE change to 4c:70mm2 XLPE(TCL-44.8kW).
\end{description}

The following changes were due to additional works detailed on drg No: EM3204.

\begin{description}
\item[MDB-BP-2BMEC]
   \begin{enumerate}
     \item Incomer MCCB 80A TP change 100A TP.
     \item FPCP-H1-2B2 cable size 4c:10mm2 XLPE change to 4c:16mm2 XLPE.
     \item FPCP-H1-2B1 cable size 4c:6mm2 XLPE change to 4c:6mm2 XLPE change to 4c:10mm2 XLPE (TCL-5.5kW).
   \end{enumerate}
\item[SMDB-FB-2BMEC]
\end{description}

The following changes were due to drawing No: EM3206.
\begin{description}
\item[MDB-BP-1B2] 
    \begin{enumerate}
       \item MDB-BP-1BCOM Additional outgoings for COM-MP-041.
       \item SMDB-BP-1BS6 MCCB 400A change to 500A and cable size 2x4c:120mm2 XLPE change to 2x4c:150mm2 XLPE (TCL-221kW).
       \item 400A TP+2x4c:120mm2 XLPE removed for FFP-3.
    \end{enumerate}
\item[SMDB-BP-1BS6] Additional outgoing for DB-LS-SR4.
\item[SMDB-BP-1BS10] Additional works were requested as follows:
    \begin{enumerate}
      \item DB-H3-1BSS2 cable size change to 2c:10mm2 XLPE change to 2c:16mm2 XLPE (TCL-1.2kW).
      \item CP-BP-1BTE/F4 cable size change to 4c:16mm2 XLPE change to 4c:25mm2 XLPE MCCB 40A TP Change to 60A TP (TCL-25kW).
      \item CP-BP-1BTF/F2 and CP-BP-1BTE/F2 MCCB 60A TP change to 80A TP (TCL-37kW).
      \item CP-BP-1BTF/F3 and CP-BP-1BTE/F3 MCCB 40A TP change to 60A TP (TCL-22kW and 25kW).
    \end{enumerate}
\end{description}







