

\chapter{Delays in Engineering, Procurement and Construction due to Frequent  Design Changes}

As per the approved Baseline programme the HLS-DSE JV were expecting to receive Good for Construction (GFC) drawings on 14 September 2013 for all areas. This would have ensured that the Contract dates could have been met. GFC drawings are an industry practice by which the Engineer signals the completion of the design and the avoidance of errors omissions and delays by using drawings such as those marked as GFE (Good for Engineering). 

The Consultant has been unable to finalize the design on time and drawings and designs were provided mostly reactively to requests and notices by the Contractor. This has subsequently caused disruption to the \JV Engineering, Submittals, Procurement and Work Progress activities.

\section{Design Changes}

Many design changes were as a response to the \JV RFIs. As of today more than 1300 RFIs have been issued. The events described below are more or less in reverse chronological order from the more recent to the earliest.

Design changes can in general be grouped in the following categories:

\begin{enumerate}
\item As responses to RFIs to resolve, space constraints. The Engineer's design was not coordinated with the basic architectural and structural design. This was most acute in the St Regis Hotel, where large beams and inadequate floor to ceiling height resulted in congested  areas. This was not evident during the tender process and has resulted in additional costs to the Contractor.

\item As responses to RFIs to provide further information due to lack of completeness of the design.

\item Errors and omissions, which either the Engineer corrected or as responses to RFIs.
\end{enumerate}

\begin{enumerate}
\item Variation notification for HVAC works was served under \letter{YL/1922 \& 1931} dated 15 March 2015 and 23 March 2015 (CAR 085). This variations to the works related to:
  \begin{enumerate}
	\item HVAC provisions for Electrical and Telecommunication rooms.
	\item Modifications to duct sizes in St. Regis Mezzanine Floor.
	\item Wow suite toilet exhaust requirements.
	\item Fan air flow changes.
	\item Updates AHUs serving F35 Westin Hotel.
	\item Modifications to duct sizes (W Hotel)
	\item Additional exhaust fan.
  \end{enumerate}
  
 \item Updated electrical drawings (38) were issued on 7 Mar 2015. We served notification as to the time impact in studying the changes (as there were no cloud revisions) and to advice impacts. Costs were reserved under CAR 146. Notifications were served via \letter{1913} and \letter{1920} dated 11 Mar 2015 and 15 Mar 2015.
 
 \item We raised concerns for missing design information for Kitchen equipment revised SLD/power drawings for W Hotel areas in order to proceed with Shop drawings preparation \letter{YL/1856} dated 5 Feb 2015.
 
 \item We raised variation notification due to additional HVAC works \letter{YL/1850} dated 5 Feb 2015.
       \begin{enumerate}
          \item BTU meters for the chilled water system.
          \item Seasonal Taste air distribution details.
          \item Relocation of AHUs at Basement-2
          \item AHU orientation, piping connection and duct re-arrangement.
          \item Additional motorized smoke dampers for standby smoke fans.
       \end{enumerate}

    
\item We requested Specialist subcontractor drawing for Data Centre and Security rooms  to proceed with MEP Shop drawings (affected Mezzanine St. Regis) \letter{YL/1813} dated 20 Jan 2015.\footnote{Security room finalization still pending as of 20 May 2015}. 

\item On 16 December 2014, we advised that MEP delays to the accelerated program were attributed to design delays, civil work delays in the casting of slabs, lack of primary co-ordination by \KA\ delays in appointing subcontractors affecting MEP works (kitchens, ELV works, IT and Audio-visual) and failure of ID designers \letter{YL/1741} dated 16 November 2014.
  
\item We received instructions to provide adaptors and sanitaryware accessories for all Hotels via \letter{Du/26882/14}  dated 16 Nov 2014. 

\item We recorded our concerns due to revisions to the design we received via sketches showing cross-contamination between fresh air intakes and exhaust from kitchens via letter ref: \letter{YL/1760} dated 24 December 2014.

\item Additional fire-fighting works for all hotels were confirmed via letter ref: \letter{YL/1765} dated 28 December 2014.

\item Revised RCP drawings were issued to the \JV which lead to abortive works at St. Regis Ground Floor to Level-2. This was confirmed by \letter{YL/1732} dated 4 December 2014.

\item We received instructions to install earthing system for all structured cabling Telecommunications rooms for all hotels (CAR 130) and \letter{YL/1728} dated 3 Dec 2014.

\item Additional UPS was added to serve the St Regis Hotel Data Center. This was confirmed by letter on 30 October 2014. (CAR 111).

\item Additional floor clean-outs were requested by the Engineer during inspections (CAR-116).

\item FCU type changed from decorative to ducted in W-hotel (CAR-115).

\item Changes to drainage services at the St Regis Hotel due to floor sinks passing over post tension slabs. (CAR 114).

\item Additional Control Panels in St. Regis Hotel Areas. (CAR-113).

\item Revisions to air outlets (CAR-117)

\item Revised Electrical works at St. Regis Hotel (CAR 83) submitted on 4/11/2014.

\item Additional Fans/AHUs that were instructed via RFI responses (CAR 85).

\item  Revised HVAC works  in St Regis Hotel (CAR 116).

\item Additional Electrical Works to St Regis Hotel Data Center (CAR 118)

\item Revised CCTV layouts for W \& Westin Hotel Areas (CAR 087)
  
\item Delays in the finalization of the Central Battery System for W \& Westin Hotels. We raised concerns via letter \letter{YL/1688}  dated 15 October 2014.

\item On 25 September 2014 we confirmed additional works for Drainage Service for one additional toilet at basement. (CAR 92).

\item On 25 September 2014, we confirmed additional works for water supply supply points for pantry at P4 in the St Regis Hotel.

\item On 25 September 2014, we confirmed additional works due to revisions to equipment schedules received via \letter{RF/MEP/M307}  (CAR 85).

\item On 25 September 2014 we confirmed additional works due to revised electrical drawings for Westin Hotel. (CAR 86).

\item On 1 October 2014 we confirmed instructions for additional electrical works due to missing power feeders at the Banquet Hall. received via RFI/MEP306 \& 309 for St Regis Hotel. (CAR 83).

\item On 2 October 2014, we confirmed additional works due to revisions of Mechanical Electrical loads. This was received as a reply to RFI/MEP/E295 for all three Hotels (CAR 83, 86 \&87). 

\item On 2 October 2014 we confirmed additional works due to revised Electrical drawings for Food and Beverage areas (CAR 93).

\item On 2 October 2014, we confirmed additional works due to Revised Mechanical Equipment schedules. This was received via RFI/MEP/M309 for all three Hotels (CAR 85).

\item On 8 October 2014, we were instructed by letter HLG/626/2.05/YE/nd/4327/14 attaching \KA letter Ref. No.: DU1211/DU/L/20844/14 dated 7 September 2004 confirming additional containment for the Access Control System.

\item On 13 October 2014, we confirmed additional works due to the addition of Electrical Heaters to some of the AHUs. This was received via RFI/MEP/M309 for all three hotels (CAR 94).

\item On 6 September 2014, we expressed concerns as to the constructibility and maintainability of the St Regis Technical 1 Plant room and issued proposals to minimize impacts. \letter{1538}. In our letter we requested \KA to resolve the issues latest within 7 days in order to minimize the impact on the accelerated target dates. Further updates were issued via letter \letter{1545} dated 9 September 2014 and \letter{1566} dated 17 September 2014 and \letter{1729} dated 4 December 2014.


\item On 7 September 2014, we requested details for the Access Control System via \letter{1542}.

\item On 9 September 2014, we expressed concerns as to changes to the Architectural design of the St. Regis Fire Pump Room which remained incomplete via \letter{1550} impeding installation of Fire Pumps.

\item On 10 September 2014, we confirmed receipt of revised HVAC drawing via \letter{1549} (CAR 85).

\item On 11 September 2014, we confirmed receipt of revised Electrical drawings for St. Regis Hotel via letter \letter{1555} (CAR 83).

\item On 22 September 2014, we issued notification for additional works related to the Domestic Cold Water system via \letter{1572} (CAR 89).

\item We received revised electrical drawings for Westin via HLG Transmittal Ref. No: HLG-626-DT-HLS-1671 dated 8 September 2014. (CAR 86)

\item We received revised electrical drawings for W Hotel via HLG Transmittal Ref. No.: HLG-626-DT-HLS-1728 dated 17 September 2014.
\end{enumerate}

The above do not record fully the method and lack of detail in issuing design information to the \JV. The general drawing issued to us did not contain adequate information to develop Shop Drawings. Clarifications and proposals were sent to the Engineer for missing information  via RFIs. 



