
\def\omar#1{\hl{#1}}

\ExplSyntaxOn
 
\edef\aprefix{ERFI}
% #1 DB #2 Suffix
%
 \NewDocumentCommand \printERFI { m m }
  {
  % sort the list in numerical order
    \SortDB { #1 }
     
 % map and print only for category (one to many also possible here)   
     \clist_map_inline:cn { #1 }
      {  
       \cs_if_exist:cT {\aprefix##1-#2}
          {
      %     \PASS RFI-E##1-#2
           \cs:w \aprefix##1-#2\cs_end:
           \index {RFI~Electrical>RFI-E-##1-#2}
           }
         % {\FAIL ##1-#2}\par
      }
  }
  
\ExplSyntaxOff    


 %#1 Number
%#2 Hotel
%#3 Impact
%#4 Description
%#5 Area

\newenvironment {ERFI} [3] {%
  \vspace*{12pt}
  \parindent0pt
  \mbox{\bfseries\color{red!80!black}\textsf{RFI-E-#1}}
        \textbf{#2} \textbf{#3} \par
  \begin{enumerate}}
  {\end{enumerate}}

% Optional DB letters W - W hotel
% WE - Westing
% SR  - St Regis

\NewDocumentCommand {\addERFI} { O{SR} +m +m +m +m +g  +g } 
{%
   \expandafter\gdef\csname \aprefix#2-#1\endcsname
   {%
      \begin{ERFI}{#2} {#3} {#4} 
        #5
     \end{ERFI}
       
      \IfNoValueTF{ #6 }{ #6 }{ } 
       
      \IfNoValueTF{ #7 }{ #7 }{ }
    } 
    \addtoDB {ERFIDB} {#2}
   \par 
}

%% Create DB 
\CreateDB {ERFIDB}   
  
\addERFI[W] {703} {10 May 2015 received 24 May 2015}{Westin and W Guestroom Floors}
{\index{W RFI>RFI-E-703}
   \item Contractor issued query based on HLG letter Ref: HLG/626/2.05/AMM/es/7764/15 dated 4 May 2015
      to request how the lighting in guestroom corridors would be controlled. The HLG letter referred to the Client instructing that lighting control in corridors be cancelled. 
   \item Engineer answered, 'grid switch to be provided in each electrical room for each floor, no of gangs to be as per circuits requirement and to proceed futher as per site conditions.  
   \hl{Omar and PMs to report time implications. Does this involve redrafting of any drawings? If yes please specify}   
}{}{}

\addERFI[WE] {703} {10 May 2015 received 24 May 2015}{Westin and W Guestroom Floors}
{
   \item Contractor issued query based on HLG letter Ref: HLG/626/2.05/AMM/es/7764/15 dated 4 May 2015
      to request how the lighting in guestroom corridors would be controlled. The HLG letter referred to the Client instructing that lighting control in corridors be cancelled. 
   \item Engineer answered, 'grid switch to be provided in each electrical room for each floor, no of gangs to be as per circuits requirement and to proceed futher as per site conditions.  
   \omar{Omar and PMs to report time implications. Does this involve redrafting of any drawings? If yes please specify}   
}{}{}


\addERFI [B] {702} {18 May 2015 received 22 Jan 2015} {St Regis, Basements 1, 2 and 3, Confirmation for lighting circuits for Executive Lift Lobby (B1) and Banquet suite Lift Lobby (B2 and B3) }%
 {
  \item Contractor observed that no lighting design was available for the Executive Lift Lobby (B1) and Banquet suite Lift Lobbies (B2 \& B3). Contractor added circuit references to the light points shown on RCP drawings which were received via \DT{2885}{} dated 5 May 2015 and \DT{2896}{} dated 10 May 2015.
  \item Co-ordinate light location/distribution with ID/Fit Out drawings. Submit RCP for review and approval. Update load schedule based on similar loads.
 }{}{}

\addERFI [W] {701} {12 May 2015 received 23 May 2015} {W Hotel Podium 2, Kitchen, missing circuit references}
 {
   \item Contractor advised of missing DB references and circuits for kitchen areas.
   \item \KA responded to be connected to nearest light DB, advisable DB-FB-2PR2 which has an available spare. \KA also requested for the load schedule to be submitted for approval.
   \omar{Omar I thought we have submitted load schedules, for W?}
 }
 {}{}

\addERFI [B] {696} {7 May 2015 received 19 May 2015} {Basement 1 Janitor Room and Staff Toilets }
{
  \item Contractor noted that the Janitor Room and Staff Toilet lighting and power design is not available.
  \item \KA responded to update the drawing and to be resubmitted for approval!
   \omar{Omar Did we submit?}   
}

\addERFI [SR] {695} {7 May 2015 received 12 May 2015 } {St Regis, PD3, PD4, PD5}
  {
    \item Contractor attached drawing, showing power supply to door requiring not shown in the Electrical design. Contractor requested feeder details for the power socket.
    \item \KA advised to connect to nearby motorized damper.
    \hl{Kyriacos, Omar, how does this affected site. This is clearly a disruptive activity.}
  }

\addERFI [W] {693} {27 Apr 2014 received 30 Apr 2014} {Westin Hotel, discrepancy between Fino RCP drawings and approved shop drawing, Podium 2, Meeting Rooms}
 {
    \item Contractor highlighted conflict between Contractor's approved Shop Drawing and Fino's latest issued RCP drawings. 
    \item \KA respondedn to follow RCP from Fino Int'l for location and number of downlights.
    \omar{Does this mean we need to go back and modify works on site?}
 }

\addERFI {692} {26 Apr 2015 received 2 May 2015 } {All levels having Sofia drawings}
  {
    \item  Contractor referred to nmerous RFI replies   and updated SLD attached.
    \item \KA responded to submit the SLD, for review assesment considering voltage drop, available NOC load etc. \omar{Did we submit? We are not responsible for Sofia loads and hence KA should not be asking us to stay with submitted Dewa loads. If these were exceeded please advise them}
  }

\addERFI [B] {691} {26 April 2015 received 28 April 2015 }{ Basement Loading and unloading area. Missing height of the isolators for scissor lifts.}
 {
   \item Contractor requested information as to the height of required scissor lift isolators.
   \item \KA instructed HLG to co-ordinate with Otis and advise.
   \omar{There are cables in the air in that vicinity, has this been resolved?}  
 }

\addERFI [W] {690} {7 May 2015 received 23 May 2015} {W Hotel, Mezzanine, discrepancy between FA design and architectural drawing.}
 {
 	\item Contractor advised conflicts between architectural drawings and FA design drawings.
 	\item \KA advised to update fire alarm drawings in co-ordination with FA Supplier and UAE code. Shop Drawing to be submitted for approval.
 }
 
\addERFI [SR] {689} {7 May 2015 received 13 May 2015}{St Regis, Podium 1 and TEchnical 1, relocation of amplifier. }
 {
	\item Contractor requested confirmation of location of amplifier rack from Podium-1 to Technical-1.
	\item \KA confirmed and noted routing of containment to be co-ordinated with other services.
	\omar{What amplifier is this? Was containment missed earlier, otherwise this is a disruptive activity.}
 } 

\addERFI [W] {688} {3 May 15 returned 18 May 2015} {W Hotel Ground Floor }
 {
	\item Contractor advised conflicts between current architectural drawings and Approved Fire Alarm Shop drawings.
	\item \KA replied to revise Shop Drawings according to UAE Fire regulations and resubmit.
	\omar {Did we resubmit?}
 }
 
 \addERFI [B] {687} {} {Basement 1 Clashing location of telephone and power sockets with lockers.}
  {
    \item We noticed that the location of power socket and telephone is clashing with locker location and requested clarifications.
    \item \KA provided Mediatech's response to move to nearest available free wall space.
    \omar {Why did we query this? Was it oicked up during an inspection?}
  }

 \addERFI [B] {686} {} {Basement 1 Fire Command and BMS Requirement}
  {
    \item Engineer made comments on \JV Shop drawing. Contractor requested clarifications.
    \item Provision for monitoring of BMS to be provided in the Fire Command Room \JV \& BMS Specialist to confirm that 1 data point and 1 socket is required only.
  }
  
\addERFI [B] {685} {28 Apr 2015 received 6 May 2015} {Basement 1 Fire Command Center - Elevator Panel Details} 
 {
 	\item Contractor requested information for the following:
 	\item \KA replied HLG shall lead the co-ordination with all subcontractors and conclude the requested information as all the submittals are approved. 
 	
 	\delay\delay\delay
 	
 	\omar{Has this now been received?}
 }
 
\addERFI [B] {684} {28 Apr 2015 received 6 May 2015} {Basement 1 Fire Command Center} 
 {
	\item Contractor requested the actual furniture layout for fire conrol room, as commented by Engineer on Shop drawings.
	\item \KA\ instructed HLG to co-ordinate with Furniture Supplier to submit proposal for Operator Approval.
	
	\omar {Zeljko, Rabih/Omar/Rahul we need to finish and get out, request furniture officially from HLG to install computers. } 
	\delay\delay\delay
 }

\addERFI [B] {683}{}{BMS}
 {
 	\item BMS DDC Panel relocation, became necessary due to space constraints.
 	\item \KA responded negatively.
 }

\addERFI [B] {682} { } {Du}
 {
 	\item \lorem
 }


 %%
\chapter{Electrical RFIs St~Regis Hotel}
\printERFI {ERFIDB} {SR}
%
\chapter {Electrical Basements}
\printERFI {ERFIDB} {B}
%
\chapter {Westin Hotel}
\printERFI {ERFIDB} {WE}
%
\chapter {W Hotel}
\printERFI {ERFIDB} {W}










