\makeatletter
\long\def\hlshadi#1{\hl{#1}}
\cxset{enumerate numberingi/.is choice,
  enumerate numberingi/.code={\renewcommand\theenumi {\csname#1\endcsname{enumi}}},
  enumerate numberingii/.code={\renewcommand\theenumii {\csname#1\endcsname{enumii}}},
  enumerate numberingiii/.code={\renewcommand\theenumiii {\csname#1\endcsname{enumiii}}},
  enumerate numberingiv/.code={\renewcommand\theenumiv {\csname#1\endcsname{enumiv}}},
  enumerate labeli punctuation/.store in=\enumeratepunctuationi@cx,
  enumerate labeli/.is choice,
  enumerate labeli/brackets/.code={\renewcommand\labelenumi{(\theenumi\enumeratepunctuationi@cx)}},
  enumerate labeli/square brackets/.code={\renewcommand\labelenumi{[\theenumi\enumeratepunctuationi@cx]}},
  enumerate labeli/right bracket/.code={\renewcommand\labelenumi{\theenumi\enumeratepunctuationi@cx)}},
  enumerate label left/.store in=\enumeratelabelleft@cx,
  enumerate label right/.code=\renewcommand\labelenumi{\enumeratelabelleft@cx\theenumi\enumeratepunctuationi@cx#1},
  enumerate leftmargini/.code={\setlength\leftmargini{#1}},
  enumerate leftmarginii/.code={\setlength\leftmarginii{#1}},
  enumerate leftmarginiii/.code={\setlength\leftmarginiii{#1}},
  enumerate leftmarginiv/.code={\setlength\leftmarginiv{#1}},
  listi topsep/.store in=\listitopsep@cx,
  listi partopsep/.store in=\listipartopsep@cx,
  listi itemsep/.store in=\listiitemsep@cx,
  listi parsep/.store in=\listiparsep@cx,
  listii topsep/.store in=\listiitopsep@cx,
  listii partopsep/.store in=\listiipartopsep@cx,
  listii itemsep/.store in=\listiiitemsep@cx,
  listii parsep/.store in=\listiiparsep@cx,
  listiii topsep/.store in=\listiiitopsep@cx,
  listiii partopsep/.store in=\listiiipartopsep@cx,
  listiii itemsep/.store in=\listiiiitemsep@cx,
  listiii parsep/.store in=\listiiiparsep@cx,
}
\cxset{compact1/.style={%
  enumerate numberingi=arabic,
  enumerate numberingii=alph,
  enumerate numberingiii=alph,
  enumerate numberingiv=roman,
  enumerate labeli punctuation=.,
  enumerate label left=,
  enumerate label right=,
  enumerate leftmargini=2.2em,
  enumerate leftmarginii=2.1em,
  enumerate leftmarginiii=1.5em,
  enumerate leftmarginiv=2em,
  listi topsep=8\p@ \@plus2\p@ \@minus\p@,
  listi itemsep=0\p@ \@plus2\p@ \@minus\p@,
  listi parsep=0\p@ \@plus2\p@ \@minus\p@,
  listii topsep=0\p@ \@plus2\p@ \@minus\p@,
  listii itemsep=0\p@ \@plus2\p@ \@minus\p@,
  listii parsep=0\p@ \@plus2\p@ \@minus\p@,
  listiii topsep=0\p@ \@plus2\p@ \@minus\p@,
  listiii itemsep=0\p@ \@plus2\p@ \@minus\p@,
  listiii parsep=0\p@ \@plus2\p@ \@minus\p@,
}}
\cxset{compact2/.style={%
  enumerate numberingi=alph,
  enumerate numberingii=roman,
  enumerate numberingiii=alph,
  enumerate numberingiv=roman,
  enumerate labeli punctuation=,
  enumerate label left=(,
  enumerate label right=),
  enumerate leftmargini=2.2em,
  enumerate leftmarginii=2.1em,
  enumerate leftmarginiii=1.5em,
  enumerate leftmarginiv=2em,
  listi topsep   = 8\p@ \@plus2\p@ \@minus\p@,
  listi itemsep = 0\p@ \@plus2\p@ \@minus\p@,
  listi parsep   = 0\p@ \@plus2\p@ \@minus\p@,
  listii topsep  = 0\p@ \@plus2\p@ \@minus\p@,
  listii itemsep= 0\p@ \@plus2\p@ \@minus\p@,
  listii parsep  = 0\p@ \@plus2\p@ \@minus\p@,
  listiii topsep = 0\p@ \@plus2\p@ \@minus\p@,
  listiii itemsep= 0\p@ \@plus2\p@ \@minus\p@,
  listiii parsep  = 0\p@ \@plus2\p@ \@minus\p@,
}}

\ExplSyntaxOn
\def\setenumerate#1{
\cxset{#1}
\def\@listi{%
           \leftmargin\leftmargini
            \parsep\listiparsep@cx
            \topsep\listitopsep@cx\relax
            \itemsep\listiitemsep@cx}
            
\def\@listii{\leftmargin\leftmarginii
            \parsep\listiiparsep@cx
            \topsep\listiitopsep@cx\relax
            \itemsep\listiiitemsep@cx}
            
\def\@listiii{\leftmargin\leftmarginiii
            \parsep\listiiiparsep@cx
            \topsep\listiiitopsep@cx\relax
            \itemsep\listiiiitemsep@cx}
}
\ExplSyntaxOff

\setenumerate{compact1}
\makeatother
\def\delay{\textcolor{red}{\Fire}}

\ExplSyntaxOn
% Holds master fields for all ODBs
\clist_new:c {DBSMASTER}

% New DBS
\clist_new:c {DBS}

%% Create the DB. The DB can have any name
%% 
\NewDocumentCommand {\CreateDB} { m }
  {
    \clist_new:c {#1}
    \clist_gput_left:cn {DBS} {#1}
  }   

% add only the number, and this only at the right,
% expecting the user to type it in ascending order and thus make sorting 
% easier
% Note the elements are stores as 1,3,4,68,112 etc.
% #1 DB name
% #2 field index - integer
% 
\NewDocumentCommand \addtoDB { m m  }
  {
    \clist_gput_right:cx { #1 } { #2 }
  }
 
%% Generate some variants
%%

\cs_generate_variant:Nn \clist_sort:Nn {cn}   

%% generalized sort DB
\NewDocumentCommand \SortDB { m }
{
  % remove any duplicates before sorting out
   \clist_remove_duplicates:c { #1 }
   % make variant here
   \clist_sort:cn {#1}
     {
       \int_compare:nNnTF { ##1 } < { ##2 }
        { \sort_reversed: }
        { \sort_ordered: }
    }
}

% #1 DB #2 Suffix
%
 \NewDocumentCommand \printRFI { m m }
  {
  % sort the list in numerical order
    \SortDB { #1 }
     
 % map and print only for category (one to many also possible here)   
     \clist_map_inline:cn { #1 }
      {  
       \cs_if_exist:cT {RFI##1-#2}
          {
            \cs:w RFI##1-#2\cs_end:
            \index {RFI~Mechanical>RFI-M-##1-#2}
          }
       
      }
  }
\ExplSyntaxOff    


 %#1 Number
%#2 Hotel
%#3 Impact
%#4 Description
%#5 Area

\newenvironment {RFI} [3] {%
  \vspace*{12pt}
  \parindent0pt
  \mbox{\bfseries\color{red!80!black}\textsf{RFI-M-#1}}
        \textbf{#2} \textbf{#3} \par
  \begin{enumerate}}
 {\end{enumerate}}

% Optional DB letters W - W hotel
% WE - Westing
% SR  - St Regis

\NewDocumentCommand {\addRFI} { O{SR} +m +m +m +m +g  +g } 
{%
   \expandafter\gdef\csname RFI#2-#1\endcsname
   {%
      \begin{RFI}{#2} {#3} {#4} 
      
        #5
      
      \end{RFI}
       
      \IfNoValueTF{ #6 }{ #6 }{ } 
       
      \IfNoValueTF{ #7 }{ #7 }{ }
    } 
    \addtoDB {MRFI} {#2}
   \par 
}

%% Create DB 
\CreateDB {MRFI}   




\chapter{Mechanical RFIs Related to St Regis}



\addRFI {497} {7 Dec 14 received 19 Jan 2015} {St Regis, Podium 1, Main Kitchen}
{
\item The RFI referred to previous response to RFI, where the Engineer directed the Contractor 
         to their comments on Shop Drawing AHC-HLS-SRH-SDM-AC-P1-0. Contractor re-iterated
         that achievable ceiling in Kitchen area could only be 2300mm and at the kitchen extract hood 2m.
\item This area was almost impossible to complete given the congestion to achieve this ceiling height.         
}


\addRFI {537} {18 Jan 2015 received 22 Jan 2015} {St Regis, Basement 1}%
 {
   \item Engineer issued instruction via drgs DU-1211-DU-00564 on 12 Jan 2014. 
    Engineer issued revised layout for AC-4020.
   \item Suggested re-arrangement blocked access to shaft 35-36/D-F.
   \item No chilled water piping and valving was not shown.
   \item Length of AHU-B2-02 is in excess of 5 meters plus a 500 mm plenum. 
 }



\addRFI{657}
{20-Apr-2015, responded 27 April 2015} {St Regis, Mezzanine, Security Room} 
{
\index{CCU>Security Room}\index{Primary co-ordination>CCTV}
\index{St Regis>Mezzanine>Security Room}

\item The layouts were not matching with latest Architectural drawings.
\item Layout not matching with HVAC Design.
        \begin{enumerate}
        \item The new layouts did not allow space for the CCUs.
        \item The new layouts did not show any access flooring.
        \end{enumerate}
\item Location of CCU 
\item Clear height not achievable.
\item Kitchen extract passing through security room, clashing with Monitor/LED display.
\item Uncertainty if there is access floor or not.
\item Engineer responded to swap store room with security room. The response did not adequately cover all the 
queries. The kitchen headroom or the access flooring was not responded to.
\item The response was followed by workshop meetings. As of May 20 2015, the Main Contractor did not carry the architectural changes required. \delay\delay\delay
}


\addRFI{662} {12-Apr-2014, responded 13-Apr-2014} {St Regis Podium 1 Winter Garden} 
{
\item This RFI dealt with incomplete issues regarding ceilings in Podium 1 Winter Garden.
\item The RFI requested details of return air grilles.
}


\addRFI {0646}{23 April 2015 responded 5 April 2015}{St Regis, Spa Area Podium 2}%
 {
 \index{St Regis>Spa>RFI-0646}
 \item Contractor requested clarifications for discrepancies between BARR \& WRAY requirement drawings and MEP design layout.
 \item Engineer responded to all queries and confirmed requirements.
 \item Area affected Spa. \delay \hl{Shadi to report on nature of delay}
 }{
 The Spa area subcontractor was appointed late. The area at Podium 2, was one of the first areas to be
 completed. The public areas around it, as well as the French Courtyard were designed very late.
 
 The Spa area subcontractor was appointed late. The area at Podium 2, was one of the first areas to be
 completed. The public areas around it, as well as the French Courtyard were designed very late.
}{
 Appointment took place only in January 15.
}

\addRFI [B] {0222}{10 Jun 2014 received 17 Jun 2014 }{Air and Dirt Separators, Boiler Room}
{
\item The Contractor requested clarification on additional air separators in the Boiler Room.
\item Relevant drawings Ref: DU1211-WS2173 Westin Hotel \& DU1211-WS2024 for Basement-1
   \begin{enumerate}
      \item Air and dirt separators are shown in B-1 Boiler plant room for the central water heating. This was not shown on Tender drawings.
      \item Air and Dirt separator is shown in Westin Hotel TE-3 plant room.
      \item Contractor requested verification and noted that the additional equipment shown would lead to additional costs to the Client.
      \item Engineer responded that they should be provided as shown on the revised drawings.
   \end{enumerate}
\item   Later Contractor made arrangements for submittals and for air-freighting later on these equipment. 
}{}{}

\addRFI [B] {0215}{2 Jun 2014 replied 12 Jun 2014} {Basement Duct clashes with Water Feature Plant Room}
{
\item Issues arose due to co-ordination of HV cables tray and ducting.
\item Engineer instructed that the duct be re-sized and re-routed.
}{}{}

\addRFI [WE] {0247} {25 June 2014  returned 5 July 2014} {Westin, Ground Floor, Insufficient ceiling void}
{
   \item Contractor advised that the ID drawings show a gap between the ceiling and the duct of 60mm and that this was going to impede return air.
   \item Please refer to attached modified ceiling levels.
}

\addRFI [B] {0655} {16 Apr 15 received 4 May 15 } {Basement 1, MEP Services in loading Bay}
{
  \item Contractor requested confirmation to Engineer's reply RFI: No. HLG-626-RFI-ME-0618 to proceed with capacity as mentioned in the RFI with one set of fans, one duty and one standby (total 2 fans). 
  \item Engineer responded to refer to RFI reply dated 21 April 2015.
}

\addRFI [W] {0661} {12 Apr 2015 received 22 Apr 2015} {W Mezzanine Floor, Male and Female Accessible Toilets RCP}
  {
    \item Contractor requested co-ordinated RCP layout with diffuser size type and location for male/female and accessible toilet to finalize Shop Drawing.
   \item Engineer responded by providing CAD and PDF files. 
   
   \hlshadi{Shadi to confirm if the drawings received were adequate. (Drawings were received by HLG Feb 15 2015.
     They seem to me they were for a Tender package. We should request Subcontractor Shop Drawings from
     Main Contractor.
    } 
  }
  
\addRFI[B] {0664}  {21 Apr 2015 received 28 Apr 2015} {Basement 1, Loading Bay Area}
 {
   \item Contractor requested clarifications and made proposals for additional fans.
   \item Engineer responded with detailed reply, including tag-number. 
 }
 
\addRFI {0666} {28 April 2015 received 14 May 2015 } {St Regis, Festival Dining Kitchen Podium~1 }
 {
   \item Contractor highlighted that the drainage for this area are not available and would cause problems in finalizing Mezzanine below.
   \item This was responded by CKP Who advised that designs have still not been received from AHG (FNB Division). These are AHG spaces (Specialty Restaurants). \delay\delay\delay
 }


\addRFI [WE] {0676} {7 May 2015 received 18 May 2015} {No access to shaft MR1 at Technical~2 Westin Hotel}
{
  \item Westin Hotel Technical 2 core wall penetration Builder's Works drawings (rev0 and rev1) we proposed 1700 x 950   opening. This was approved by the Engineer. This size has been revised due to the comment, changes were then incorporated in rev 1 layout. At Site the available opening is 2440 x 600. Contractor included an Annexure with proposals.
  \item Engineer responded with proposal to reduce space further between pipes. \hl{Shadi to confirm what eventually was followed.}
}

\printRFI {MRFI} {SR} 

%% Typeset all Chapters
%% 
\chapter {Mechanical RFIs Related to Westin Hotel}
\printRFI {MRFI} {WE}


\chapter{Mechanical RFIs Related to `W'~Hotel}
\printRFI {MRFI} {W}

\chapter{Mechanical RFIs Related to `Basements'  }
\printRFI {MRFI} {B}
  


