\cxset{image=greek-men}

\parindent1em

\chapter{Greek}
\epigraph{The Pleiads have left the sky, and\\
the moon has vanished. It’s midnight:\\
the time for meeting is over.\\
And me—I am lying, lonely}{Sappho}
\label{s:greek}
\index{languages>Greek}\index{Herodotus}\index{alphabets>Greek}



\enquote{The Phoenicians who came with Kadmos,} wrote Herodotus in the fifth century BC of the legendary Phoenician prince of Tyre and brother of Europa, ``\ldots introduced into Greece, after their settlement in the country, a number of accomplishments of which the most important was writing, an art which probably was unknown to the Greeks until then''. 

A basis for the remarkable history of the Greek language is the invention of the Greek alphabet. It was modelled after Semitic scripts, with the important improvement that not only consonants but also vowels are represented by independent
letters.\footnote{Cover image, from \href{http://www.pappaspost.com/todays-undesirable-muslims-were-yesteryears-greeks-pure-american-no-rats-no-greeks/}{papaspost.com}, showing Greek immigrants arriving at Ellis Island in 1911.}

The poet Sappho had access to an alphabetic script, invented
for the Greek language just a couple of hundred years before her
time. The Greek alphabet is very similar to the Latin one, which
is the one used for English. In fact, the Latin alphabet is derived
from a variant of the Greek one.The similarity is easy to observe.
Here is the original poem, written in the Greek alphabet:\footcite{janson:2002}


\begin{center}
\arial 
ΔΕΔΗΚΕ ΜΕΝ Α ΣΕΛΑΝΝΑ\\
ΚΑΙ ΠΛΕΙΑΔΕΣ. ΜΕΣΑΙ ΔΕ\\
ΝΥΚΤΕΣ. ΠΑΡΑ Δ'ΕΡΧΕΤ'ΩΡΑ.\\
ΕΓΩ ΔΕ ΜΟΝΑ ΚΑΤΕΥΔΩ\\
\end{center} 

Transcribed in the Latin alphabet:

\begin{center}
DEDUKE MEN A SELANNA\\
KAI PLEIADES. MESAI DE\\
NUKTES. PARA D’ ERKHET’ ORA.\\
EGO DE MONA KATEUDO.\\
\end{center}

The Greek alphabet is the script that has been used to write the Greek language since the 8th century BC. It was derived from the earlier Phoenician alphabet, and was in turn the ancestor of numerous other European and Middle Eastern scripts, including Cyrillic and Latin.[3] Apart from its use in writing the Greek language, both in its ancient and its modern forms, the Greek alphabet today also serves as a source of technical symbols and labels in many domains of mathematics, science and other fields.

In its classical and modern forms, the alphabet has 24 letters, ordered from alpha to omega. The alphabet went through many stages of development, with local states using variants of the letters. The standard work on local scripts is still Jeffrey's.\footcite{jeffery1963}







Like Latin and Cyrillic, Greek originally had only a single form of each letter; it developed the letter case distinction between upper-case and lower-case forms in parallel with Latin during the modern era.

\tex has built-in commands for the usage of the Greek alphabet see section \ref{greek} in the Symbols chapter.

\bgroup
\obeylines
\greek\obeyspaces

Α	ἄλφα	aleph	alpha	[alpʰa]	[ˈalfa]	Listeni/ˈælfə/
Β	βῆτα	beth	beta	[bɛːta]	[ˈvita]	/ˈbiːtə/, US /ˈbeɪtə/
Γ	γάμμα	gimel	gamma	[ɡamma]	[ˈɣama]	/ˈɡæmə/
Δ	δέλτα	daleth	delta	[delta]	[ˈðelta]	/ˈdɛltə/
Η	ἦτα	  heth	   eta	 [hɛːta], [ɛːta]	[ˈita]	/ˈiːtə/, US /ˈeɪtə/
Θ	θῆτα	teth	theta	[tʰɛːta]	[ˈθita]	/ˈθiːtə/, US Listeni/ˈθeɪtə/
Ι	ἰῶτα	yodh	iota	[iɔːta]	[ˈʝota]	Listeni/aɪˈoʊtə/
Κ	κάππα	kaph	kappa	[kappa]	[ˈkapa]	Listeni/ˈkæpə/
Λ	λάμβδα	lamedh	lambda	[lambda]	[ˈlamða]	Listeni/ˈlæmdə/
Μ	μῦ	mem	mu	[myː]	[mi]	Listeni/ˈmjuː/; occasionally US /ˈmuː/
Ν	νῦ	nun	nu	[nyː]	[ni]	/ˈnjuː/ (US /ˈnuː/)
Ρ	ῥῶ	reš	rho	[rɔː]	[ro]	Listeni/ˈroʊ/
Τ	ταῦ	taw	tau	[tau]	[taf]	/ˈtaʊ/ or /ˈtɔː/
\egroup

With a suitable font such as |Arial Unicode MS| you do not need to do anything special to typeset short paragraphs of Greek text. Just use any editor set to encode the text in \utfviii. The example below was just cut and pasted. If you are going to write extensively in Greek it would be preferable to get a virtual keyboard. If you are using windows these are pre-build. 

\topline
\begin{quote}
Ἡροδότου Ἁλικαρνησσέος ἱστορίης ἀπόδεξις ἥδε, ὡς μήτε τὰ γενόμενα ἐξ ἀνθρώπων τῷ χρόνῳ ἐξίτηλα γένηται, μήτε ἔργα μεγάλα τε καὶ θωμαστά, τὰ μὲν Ἕλλησι, τὰ δὲ βαρβάροισι ἀποδεχθέντα, ἀκλεᾶ γένηται, τὰ τε ἄλλα καὶ δι' ἣν αἰτίην ἐπολέμησαν ἀλλήλοισι.[2]

Herodotus of Halicarnassus, his Researches are set down to preserve the memory of the past by putting on record the astonishing achievements of both the Greeks and the Barbarians; and more particularly, to show how they came into conflict.[3]
\end{quote}
\bottomline

{\small
\begin{scriptexample}{greek}
\unicodetable{greek}{% 
"0370,"0380,"0390,"03A0,"03B0,"03C0,"03D0,"03E0,"03F0}
\end{scriptexample}
}

\subsection{Greek diacritics}
\index{Greek>polytonic}

The ancient Greek writing included for many accents and diacritics. The extended unicode standard provides slots for all diacritics. Greek orthography has used a variety of diacritics starting in the Hellenistic period. The complex polytonic orthography notates Ancient Greek phonology. The simple monotonic orthography, introduced in 1982, corresponds to Modern Greek phonology, and requires only two diacritics.

Polytonic orthography (πολύς "much", "many", τόνος "accent") is the standard system for Ancient Greek. The acute accent ( ´ ), the grave accent ( ` ), and the circumflex ( ῀ ) indicate different kinds of pitch accent. The rough breathing ( ῾ ) indicates the presence of an /h/ sound before a letter, while the smooth breathing ( ᾿ ) indicates the absence of /h/.

Since in Modern Greek the pitch accent was replaced by a dynamic accent, and the /h/ was lost, most polytonic diacritics have no phonetic significance, and merely reveal the underlying Ancient Greek etymology.

Monotonic orthography (μόνος "single", τόνος "accent") is the standard system for Modern Greek. It retains a single accent or tonos (΄) to indicate stress and the diaeresis (¨) to indicate a diphthong: compare modern Greek παϊδάκια /pajˈðaca/ "lamb chops", with a diphthong, and παιδάκια /peˈðacia/ "little children" with a simple vowel. Tonos and diaeresis can be combined on a single vowel, as in the verb ταΐζω (/taˈizo/ "to feed").

\medskip
\begin{scriptexample}[]{Greek}
\unicodetable{greek}{%
"1F00,"1F10,"1F20}
\end{scriptexample}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    Greek Language
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\input{./languages/philokalia}

\section{Greek-derived scripts}

Because of Greece’s military (Alexander the Great), economic
and cultural influence, the Greek alphabet became the
prototype for the ‘complete’ (that is, fully vowelized) alphabets
that emerged in Europe in the following centuries. These eventually
diffused, almost exclusively through Greek’s granddaughter
alphabets Latin and Cyrillic, throughout the entire
world --- a process still going on over two thousand years later
(illus. \ref{fig:greekderived}).

In first-millenium BC Asia Minor (today’s Turkey), the
Greek alphabet inspired an impressive number of non-Greek
peoples to elaborate their own Anatolian alphabets: \nameref{carian},
\nameref{sec:lydian}, \nameref{sec:lycian}, Pamphylian, Phrygian, Pisidian (of the Roman
period) and Sidetic.  Nonetheless, these scripts failed to
acquire lasting significance because of the region’s declining
economic fortunes followed by several major invasions.

\input{./languages/greek-forest}

The Armenian monk St Mesrob (c. 345–440) is said to have
elaborated the Armenians’ first script c. AD 405 – Armenian is a
separate branch of the Indo-European superfamily of languages
(to which Greek and Germanic, which includes English, also
belong). Based on the Greek alphabet, the Armenian script
originally consisted of around 36 mainly capital letters. By the
1200s, Armenian notrgir, or cursive writing, had been developed,
then replacing writing in capitals (illus. 100).

St Mesrob is also credited with devising the Georgian alphabet
in the early 400s AD – Georgian is a Caucasian, not an Indo-
European, language – as well as the Albanian alphabet. (Such
multiple attributions suggest that Mesrob’s role was apocryphal.)

The ecclesiastical Georgian script used 38 letters; over
time, several styles of writing Georgian developed, with varying
numbers of letters (illus. 101). The mkhedruli, or ‘lay hand’,
which began as a medium for non-sacral texts, is Georgian’s
most frequently employed script, still in use today.

In Egypt, the Greek alphabet inspired the Coptic
alphabet that replaced one of the world’s oldest writing traditions.
In the Balkans, Greek generated the Glagolitic and
Cyrillic scripts, which eventually generated the Russian script,




\greektxt{πεῖραι} accented P has circumflex because -αι is final
and counts as short

\greektxt{πείραις} accented P has acute because αι is not final\\
and counts as long

\greektxt{ἄνθρωποι} acute on A with short U since -οι is final\\
and counts as short

\greektxt{ἀνθρώποις} acute on P with long U since οι is not final
and counts as long


The \textbf{proclitics} are\\
a. the article forms 

