\cxset{palette budapest}

\chapter{ltlists.dtx}
\label{kernel:ltlists}
 \section{List, and related environments}

This section of the kernel is interesting in order to understand the underlying general design
decisions of the \latexe Team.

The kernel source can be considered as a framework, upon which classes build. All \latex lists are aware
of their environment and can be included in other lists. You can also even if you want include sections within
enumerated or itemized lists and they will not break. You can include boxes or minipages. 

This section of the kernel creates generic factory commands for the creation of list environments. 
These list environments \refEnv{enumerate}, \refEnv{itemize} are defined in the kernel, whereas others are normally
defined in the class files (see \vref{ch:bookclass}, \vref{sec:clo} and \vref{sec:booklists}).

To understand how lists are created we need to understand that they are shaped paragraphs. Lamport's 
decision to shape them with \cs{parshape}, has produced a flexible system, used later on by standard classes to create numerous environments. The complexity of the code is necessary to make sure
that at every item the |\parshape| is used and also in any following paragraphs. Remember
that |\parshape| will only act on one paragraph. As Knuth said the \latexe Team did some
nifty tricks with |\parshape|.

A sideline is that the code can be used to add vertical and bottom space to content as well as adjust the left and right margins of text blocks. In the \latexe classes they are used to define the environments for quotations, quotes \vref{sec:quotationenvironment}, verse \autoref{sec:verseenvironment}  and verbatim environments.


\section{Generic Command}

The |\list| is the generic command that is used to create other lists.  Think of it as the |\newcommand|
for new list environments. The command takes two arguments.


\begin{quote}
        |\list|\marg{label}\marg{commands} ... |\endlist|
\end{quote}

 which can be invoked by the user as the |list| environment.  The \meta{label}
 argument specifies item labeling.  \meta{commands} contains commands for
 changing the horizontal and vertical spacing parameters.

 Each item of the environment is begun by the command
 |\item[|item label|]|
 which produces an item labeled by |itemlabel|.  If the argument is
 missing, then the |label| argument of the |\list| command is used as the
 item label. This enables the user to type |item|  or |item[some label]|.

 The label is formed by putting |\makelabel|\marg{ITEMLABEL} in an hbox
 whose width is either its natural width or else |\labelwidth|,
 whichever is larger.  The |\list| command defines \refCom{makelabel} to have
 the default  definition:
 \begin{quote}
     |\makelabel|\marg{arg} == BEGIN |\hfil| arg END
 \end{quote}
 which, for a label of width less than \refCom{labelwidth}, puts the label
 flushright, \refCom{labelsep} to the left of the item's text.  However,
 |\makelabel| can be |\let| to another command by the \refCom{list}'s
\meta{commands} argument.

 A \refCom{usecounter}\marg{foo} command in the second argument causes the
 counter \emph{foo} to be initialized to zero, and stepped by every
 |\item| command without an argument.  (|\label| commands within the
 list refer to this counter.)

 When you leave a list environment, returning either to an enclosing
 list or normal text mode, \latexe begins a new paragraph if and only if
 you leave a blank line after the |\end| command.  This is accomplished
 by the \refCom{@endparenv} command.

 Blank lines are ignored every other reasonable place--i.e.:
 \begin{itemize}
  \item  Between the |\begin{list}| and the first |\item|,
  \item  Between the |\item| and the text of that item.
  \item Between the end of the last item and the |\end{list}|.
 \end{itemize}

 For an environment like \docAuxEnvironment{quotation}, in which items are not labeled,
 the entire environment is a single item.  It is defined by
 letting |\quotation| == |\list{}{...}\item\relax|.  (Note the
 |\relax|, there in case the first character in the environment is a
 '['.)  The spacing parameters provide a great deal of flexability in
 designing the format, including the ability to let the indentation of
 the first paragraph be different from that of the subsequent ones.

 The trivlist environment is equivalent to a list environment
 whose second argument sets the following parameter values:
 \begin{description}
 \item[{\refCom{leftmargin}}] = 0: causes no indentation of left margin
 \item[\refCom{labelwidth} = 0:] see below for precise effect this has.
 \item[\refCom{itemindent} = 0:] with a null label, makes first paragraph
        have no indentation.  Succeeding paragraphs have |\parindent|
        indentation.  To give first paragraph same indentation, set
        |\itemindent| = |\parindent| before the |\item[]|.
 \end{description}

 Every |\item| in a trivlist environment must have an argument---in
 many cases, this will be the null argument (|\item[]|).  The trivlist
 environment is mainly used for paragraphing environments, like
 verbatim, in which there is no margin change.  It provides the same
 vertical spacing as the list environment, and works reasonably well
 when it occurs immediately after an |\item| command in an enclosing
 list.

\begin{texexample}{understanding how lists are being build}{ex:parshape1}

\makeatletter
\def\mybullet{$\bullet$}
\leftskip0pt \rightskip0pt

\newlength\lmargin
\newlength\rmargin
\newlength\lnwidth
\setlength\lnwidth{\linewidth}

\setlength\lmargin{30pt}
\setlength\rmargin{30pt}
\setlength\lnwidth{\textwidth}
\advance\lnwidth-2\rmargin\relax %

\hskip-\lmargin\hbox to \lmargin{\hfill\mybullet}%
\parshape 1 \dimexpr\lmargin\relax \lnwidth\relax
\lorem

% Second list level
\addtolength\lmargin{30pt}\relax
\hskip-\lmargin\hbox to \lmargin{\hfill\mybullet}%
\parshape 1 \dimexpr\lmargin\relax \dimexpr\linewidth-3\rmargin\relax
\lorem

% Third list level
\addtolength\lmargin{30pt}\relax
\hskip-\lmargin\hbox to \lmargin{\hfill\mybullet}%
\parshape 1 \dimexpr\lmargin\relax \dimexpr\linewidth-5\rmargin\relax
\lorem

\makeatother
\end{texexample}

The above were manually defined so that you can understand how the code builds the list. These can be abstracted into a macro similar to |item| to typeset the text. The label can be typeset with a new macro |mklabel|. We will do this first and change it to an enumerated list. I will use a counter that is already defined for this book and use a macro |\inc| to step it up.

\emphasize{mybullet,itm}
\begin{texexample}{understanding how lists are being build}{ex:parshape2}
\makeatletter
\def\mybullet{{\itshape\large\inc. }}
\leftskip0pt \rightskip0pt
\setlength\lmargin{0pt}
\setlength\rmargin{0pt}

\def\itm{%
  \setlength\lmargin{0pt}
  \setlength\rmargin{0pt}
  \hskip-\lmargin\hbox to \lmargin{\hfill\mybullet}%
  \parshape 1 \dimexpr\lmargin\relax \dimexpr\linewidth-\rmargin\relax}

% First level
\itm \lorem

% Second list level
\itm \lorem

% Third list level
\itm \lorem

\makeatother
\end{texexample}

Remember that \latexe defines all lengths to be global lengths, so we didn't have to redeclare them above, just reset them to the values we needed them.


 \subsection{List and Trivlist}

 The following variables are used inside a list environment:
 \begin{description}
 \item[] \refCom{@totalleftmargin} The distance that the prevailing left
     margin is indented from the outermost left margin,
 \item[] \refCom{linewidth} The width of the current line.  Must be
     initialized to |\hsize|.
 \item[] \refCom{@listdepth} A count for holding current list nesting depth.
 
 \item[] \refCom{makelabel} A macro with a single argument, used to
   generate the label from the argument (given or implied)
   of the |\item| command. Initialized to |\@mklab| by the |\list|
   command.  This command must produce  some stretch---i.e., an
   |\hfil|.
   
 \item[] \refCom{if@inlabel} A switch that is false except between the time
   an |\item| is encountered and the time that \TeX{}
   actually enters horizontal mode.  Should be tested by commands
   that can be messed up by the list environment's use of |\everypar|.
   
 \item[] \refCom{@labels} When |@inlabel = true|, it holds the labels
   to be put out by |\everypar|.
   
 \item[] \refCom{if@noparitem} A switch set by |\list| when
   |@inlabel = true|.
      Handles the case of a |\list| being the first thing in an item.
      
 \item[] \refCom{if@noparlist} A switch set true for a list that begins an
   item.  No |\topsep| space is added before or after |\item|'s such a
   list.
   
 \item[] \refCom{if@newlist} Set true by |\list|, set false by the first
   text (by |\everypar|).
   
 \item[]  \refCom{if@noitemarg} Set true when executing an |\item| with no
   explicit argument.  Used to save space. To save time, make two
   separate  \refCom{@item} commands.
   
 \item[] \refCom{if@nmbrlist} Set true by |\usecounter| command, causes
   list to be numbered.
   
 \item[] |noskipsec| A switch set true by a sectioning command when
    it is creating an in-text heading with |\everypar|.
 \end{description}


 Throughout a list environment, |\hsize| is the width of the current
 line, measured from the outermost left margin to the outermost right
 margin.  Environments like tabbing should use \refCom{linewidth} instead of
 |\hsize|.

 Here are the parameters of a list that can be set by commands in
 the |\list|'s \meta{commands} argument.  These parameters are all TeX
 skips or dimensions (defined by |\newskip| or |\newdimen|), so the
 usual \TeX\ or \LaTeX\ commands can be used to set them.  The
 commands will be executed in vmode if and only if the |\list| was
 preceded by a |\par| (or something like an |\end{list}|), so the
 spacing parameters can be set according to whether the list is
 inside a paragraph or is its own paragraph.


 \paragraph{Vertical Spacing (skips)}
 
 \begin{description}
 \item[\refCom{topsep}] Space between first item and preceding paragraph.
 \item[\refCom{partopsep}] Extra space added to \refCom{topsep} when
        environment starts a new paragraph (is called in vmode).
 \item[\refCom{itemsep}] Space between successive items.
 \item[\refCom{parsep}] Space between paragraphs within an item -- the
                 |parskip| for this environment.
 \end{description}

 \paragraph{Penalties}
 
 \begin{description}

 \item[] \refCom{@beginparpenalty} put at the beginning of a list
 \item[] \refCom{@endparpenalty} put at end of list
  \item[] \refCom{@itempenalty} put between items.
  \end{description}

 \subsection{Horizontal Spacing (dimens)}
 \begin{description}
 \item[]  \refCom{leftmargin} space between left margin of enclosing
   environment (or of page if top level list) and left margin of
                     this list.  Must be nonnegative.
  \item[] \refCom{rightmargin} analogous.
  \item[] \refCom{listparindent} extra indentation at beginning of every
     paragraph of a list except the one started by the \refCom{item}
                      command.  May be negative!  Usually, labeled
                       lists have \refCom{listparindent} equal to zero.
   \item[]  \refCom{itemindent} extra indentation added right BEFORE an item
                      label.
  \item[] \refCom{labelwidth} nominal width of box that contains the label.
                      If the natural width of the
                         label $< =$ \refCom{labelwidth},
                      then the label is flushed right inside a box
                      of width \refCom{labelwidth} (with an |\hfil|).
                      Otherwise,
                      a box of the natural width is employed, which
                       causes an indentation of the text on that line.
     \item[] \refCom{labelsep} space between end of label box and text of
                      first item.
  \end{description}


 \subsection{Default Values}
 
 Defaults for the list environment are set as follows.
 First, \refCom{rightmargin}, \refCom{listparindent} and \refCom{itemindent}
 are set
      to 0pt.  Then, one of the commands
      \refCom{@listi}, |@listii|,\ldots,|@listvi|
      is called, depending upon the current level of the list.
      The |@list|\ldots commands should be defined by the document
      style.  A convention that the document style should follow is
      to set \refCom{leftmargin} to
      |leftmargini|,\ldots, |leftmarginiv| for
      the appropriate level.  Items that aren't changed may be left
      alone, but everything that could possibly be changed must be
      reset.


\LinesNumbered
\begin{algorithm}
\caption{The \refCom{list} environment}
  \refCom{list}\marg{LABEL}\marg{COMMANDS} ==\\
   \Begin{
     \eIf{\refCom{@listdepth} > 5}{
        LaTeX error: `Too deeply nested'}{
        \refCom{@listdepth} :=G \refCom{@listdepth} + 1\\
     }
     \refCom{rightmargin}     := 0pt\\
     \refCom{listparindent}   := 0pt\\
     \refCom{itemindent}      := 0pt\\
     eval(@list \refCom{romannumeral}\refCom{the}\refCom{@listdepth})\\  
     \refCom{@itemlabel}      :=L LABEL\\
     \refCom{makelabel}       == \refCom{@mklab}\\
     @nmbrlist :=L false\\
     SETTING COMMANDS\\
     \refCom{@trivlist}\\  
     parskip          :=L \refCom{parsep}\\
     parindent        :=L \refCom{listparindent}\\
     \refCom{linewidth}        :=L \refCom{linewidth} - \refCom{rightmargin} -\refCom{leftmargin}\\
     \refCom{@totalleftmargin} :=L \refCom{@totalleftmargin} + \refCom{leftmargin}\\
     parshape 1 \refCom{@totalleftmargin} \refCom{linewidth}\\
     \cs{ignorespaces} 
   }
\end{algorithm}

The \refCom{endlist} simply adjusts the listdepth counter and ends the \refCom{trivlist}.

\begin{algorithm}
 \refCom{endlist} == \\
  \Begin{
    \refCom{@listdepth} :=G \refCom{@listdepth} -1\\
    \refCom{endtrivlist}\\
  }
\end{algorithm}

The \refCom{@trivlist} is define as,

\begin{algorithm}
 \refCom{@trivlist} ==\\
  \Begin{
    \If{@newlist = T}{\refCom{@noitemerr}}
     This command removed for some forgotten reason.\\
     \refCom{@topsepadd} :=L \refCom{topsep}\\
     \If{@noskipsec}{leave vertical mode}
     \eIf{vertical mode}{
        \refCom{@topsepadd} :=L \refCom{@topsepadd} + \refCom{partopsep}}{
        \cs{unskip} \cs{par}}
     \eIf{@inlabel = true}{
         @noparitem :=L true
         @noparlist :=L true}{
         @noparlist :=L false
             \refCom{@topsep}   :=L \refCom{@topsepadd}}
     \refCom{@topsep}      :=L \refCom{@topsep} + \refCom{parskip}\\
      Restore paragraphing parameters\\
     \cs{leftskip}     :=L 0pt\\  
     \cs{rightskip}    :=L \refCom{@rightskip}\\
     \cs{parfillskip}     :=L 0pt + 1fil\\
   NOTE: \cs{@setpar} called on every \refCom{list} in case \cs{par} has been\\
   temporarily  munged before the \refCom{list} command.\\
     \cs{@setpar}{if @newlist = false then {\cs{@@par}} fi}\\
     \refCmd{@newlist}         :=G T\\
     \refCmd{@outerparskip}    :=L \cs{parskip}\\
 }
\end{algorithm}

\begin{algorithm}
 \refCom{trivlist}  ==\\
 \Begin{
  \refCom{parsep} := \cs{parskip}\\
  \cs{@nmbrlist} := false\\
  \refCom{@trivlist}\\
  \refCom{labelwidth} := 0\\
  \refCom{leftmargin} := 0\\
  \refCom{itemindent} := \refCom{parindent}\\
  \refCom{@itemlabel} :=L "empty"\\ 
  \refCom{makelabel}\marg{LABEL} == LABEL\\
 }
\end{algorithm}



\begin{algorithm}
 \refCom{endtrivlist} ==\\
 \Begin{
     \If{@inlabel = T}{\refCom{indent}}
     \If{horizontal mode}{\refCom{unskip} \refCom{par}}
     \eIf{@noparlist = true}{}{
        \If{\refCom{lastskip} > 0}{
              \refCom{@tempskipa} := \refCom{lastskip}
              \refCom{vskip} - \refCom{lastskip}
              \refCom{vskip} \refCom{@tempskipa} -\refCom{@outerparskip} + \refCom{parskip}
             }
           \refCom{@endparenv}
     }
   }
\end{algorithm}


\begin{algorithm}
 \refCom{@endparenv} ==
   \Begin{
    \refCom{addpenalty}\marg{@endparpenalty}\\
    \refCom{addvspace}\marg{\refCom{@topsepadd}}\\
     ends the \refCom{begin} command's \refCom{begingroup}\\
    \cs{endgroup}\\
    \cs{par}  ==  \Begin{%
                  \refCom{@restorepar}\\
                  \cs{everypar}|{}|
                  \cs{par}
                  }
    \refCom{everypar} == \Begin{remove \refCom{lastbox} \refCom{everypar}|{}|}
   to match the \refCom{end} commands \refCom{endgroup}\\
    \refCom{begingroup}  
   }
\end{algorithm}

\index{kernel>lists>\textbackslash item}
The definition of item, is fairly simple deferring the complexity
to \refCom{@item} which follows.

\begin{algorithm}[htbp]
 \refCom{item} == \Begin{
    \If{math mode}{issue warning}
    \eIf{ next char = {\tt [}}{
             \refCom{@item}}{
              @noitemarg := true\\
             \refCom{@item}[\refCom{@itemlabel}]}
  }
\caption{The algorithm for \textbackslash item}
\end{algorithm}


\begin{algorithm}
 \refCom{@item}[LAB] ==
    \Begin{
     \eIf{@noparitem = true}{
        @noparitem := false
             % NOTE: then clause  hardly every taken,\\
             %  so made a macro \refCom{@donoparitem}\\
            \refCom{box}\refCom{@labels} :=G\\
             \refCom{hbox} \Begin{hskip -\refCom{leftmargin}\\
                                   {box}   \\refCom{@labels}\\
                                   {hskip} \refCom{leftmargin}}
            \If{@minipage = false}{
               \refCom{@tempskipa} := \refCom{lastskip}\\
               \refCom{vskip} -\refCom{lastskip}\\
               \refCom{vskip} \refCom{@tempskipa} + \refCom{@outerparskip} - \refCom{parskip}}
            }{
          \If{@inlabel = true}{
              then \refCom{indent} \refCom{par}   % previous item empty.
           }
           \If{hmode}{then 2 \refCom{unskip}'s\\
                           % To remove any space at end of prev.\\
                           % paragraph that could cause a blank line.\\
                     \refCom{par}\\
           }
           \eIf{if @newlist = T}{
                \eIf{@nobreak = T}{ 
                      % Kludge if list follows \refCom{section}\\
                      \refCom{addvspace}\marg{\refCom{@outerparskip} - \refCom{parskip}}}{
                       \refCom{addpenalty}\marg{\refCom{@beginparpenalty}}\\
                       \refCom{addvspace}\marg{\refCom{@topsep}}\\
                       \refCom{addvspace}\marg{-\refCom{parskip}}\\  
                    }}{
                \refCom{addpenalty}\marg{\refCom{@itempenalty}}\\
                \refCom{addvspace}\marg{\refCom{itemsep}}\\
            }

            @inlabel :=G true\\
     }

     \refCom{everypar}\{ @minipage :=G F\\
                @newlist :=G F\\
                \If{@inlabel = true}{
                    @inlabel :=G false
                       \refCom{hskip} -\refCom{parindent}\\
                       \refCom{box}\refCom{@labels}\\
                       \refCom{penalty} 0\\
                       \refCom{box}\refCom{@labels} :=G null\\
                }
                \refCom{everypar}\{\} \}\\
     @nobreak :=G false\\
     \If{@noitemarg = true}{
        @noitemarg := false\\
            \If{@nmbrlist}{
               \refCom{refstepcounter}\{\refCom{@listctr}\}
            }
     }
     \refCom{@tempboxa}   :=L \refCom{hbox}\marg{\refCom{makelabel}\marg{LAB}}\\
     \refCom{box}\refCom{@labels} :=G \refCom{@labels}\refCom{hskip}\refCom{itemindent}\\
                       \refCom{hskip} - (\refCom{labelwidth} + \refCom{labelsep})\\
                 \eIf{\refCom{wd}\refCom{@tempboxa} > \refCom{labelwidth}}{
                          \refCom{box}\refCom{@tempboxa}}{
                          \refCom{hbox} to \refCom{labelwidth}\{\refCom{unhbox}\refCom{@tempboxa}\}\\
                 } 
      \refCom{hskip}\refCom{labelsep}\\
     \refCom{ignorespaces} %gobble space up to text
  }
\end{algorithm}




\begin{algorithm}
 \refCom{makelabel}\marg{LABEL} == ERROR\\
    default to catch lonely \refCom{item}

 \refCom{usecounter}\marg{CTR} == \Begin{
                              @nmbrlist :=L true\\
                              @listctr == CTR\\
                              setcounter\marg{CTR}\marg{0}
                             }
\end{algorithm}



 

\begin{docCommand}{topsep} {\meta{dim}}
The space between the first item and the preceding paragraph.
\end{docCommand}

\begin{docCommand}{partopsep} {\meta{dim}}
The space between the first item and the preceding paragraph.
\end{docCommand}

\begin{docCommand}{itemsep} {\meta{dim}}
The space between the first item and the preceding paragraph.
\end{docCommand}

\begin{docCommand}{@itemsep} {\meta{dim}}
The space between the first item and the preceding paragraph.
\end{docCommand}

\begin{docCommand}{@topsepadd} {\meta{dim}}
The space between the first item and the preceding paragraph.
\end{docCommand}

\begin{docCommand}{outerparskip} {\meta{dim}}
The space between the first item and the preceding paragraph.
\end{docCommand}

\begin{docCommand}{parsep} {\meta{dim}}
The space between the first item and the preceding paragraph.
\end{docCommand}

The skip registers \docAuxCommand{topskip}, \docAuxCommand*{partopsep}, 
\docAuxCommand{itemsep}, \docAuxCommand{parsep}, \docAuxCommand{@topsep},
\docAuxCommand{@topsepadd}, \docAuxCommand{outerparskip} are created.

    \begin{teX}
\newskip\partopsep
\newskip\itemsep
\newskip\parsep
\newskip\@topsep
\newskip\@topsepadd
\newskip\@outerparskip
    \end{teX}
 
 
\begin{docCommand}{leftmargin} {\meta{dim}}
|\newdimen\leftmargin|
\end{docCommand}

\begin{docCommand}{rightmargin} {\meta{dim}}
|\newdimen\rightmargin|

\end{docCommand}

\begin{docCommand}{listparindent} {\meta{dim}}
|\newdimen\listparindent|
\end{docCommand}

\begin{docCommand}{labelwidth} {\meta{dim}}
|\newdimen\labelwidth|
\end{docCommand}


\begin{docCommand}{labelsep} {\meta{dim}}
 |\newdimen\labelsep|
\end{docCommand}

\begin{docCommand}{linewidth} {\meta{dim}}
|\newdimen\linewidth|
\end{docCommand}

\begin{docCommand}{itemindent} {\meta{dim}}
|\newdimen\itemindent|
\end{docCommand}

\begin{docCommand}{@totalleftmargin} {\meta{dim}}
|\newdimen\@totalleftmargin|  |\@totalleftmargin=\z@|
\end{docCommand}


 
\begin{docCommand} {leftmargini} { \meta{dim}}
\end{docCommand}
\begin{teX}
\newdimen\leftmargini
\end{teX}
\begin{docCommand} {leftmarginii} { \meta{dim}}
\end{docCommand}
\begin{teX}
\newdimen\leftmarginii
\end{teX}

\begin{docCommand} {leftmarginiii} { \meta{dim}}
\end{docCommand}
\begin{teX}
\newdimen\leftmarginiii
\end{teX}

\begin{docCommand} {leftmarginiv} { \meta{dim}}
\end{docCommand}
\begin{teX}
\newdimen\leftmarginiv
\end{teX}
    \begin{teX}
\newdimen\leftmarginv
    \end{teX}


\begin{docCommand}{@listdepth}{}
\end{docCommand}
\begin{docCommand}{@itempenalty}{}
\end{docCommand}
\begin{docCommand}{@beginparpenalty}{}
\end{docCommand}
\begin{docCommand}{@endparpenalty}{\meta{penalty}}
|\newcount\@endparpenalty|
\end{docCommand}
    \begin{teX}
\newcount\@listdepth \@listdepth=0
\newcount\@itempenalty
\newcount\@beginparpenalty

    \end{teX}


 \begin{docCommand}{@labels} { \meta{box} }
 \end{docCommand}
    \begin{teX}
\newbox\@labels
    \end{teX}
 


 \begin{docCommand}{if@inlabel}{}
   \docAuxCommand{@inlabel}true and false 
  \end{docCommand}
    \begin{teX}
\newif\if@inlabel \@inlabelfalse
    \end{teX}



\begin{docCommand}{ if@newlist} { }
\end{docCommand}
  \begin{teX}
\newif\if@newlist   \@newlistfalse
    \end{teX}
  
  \begin{docCommand}{if@newlistfalse }{}
  \end{docCommand}
  \begin{docCommand}{if@newlisttrue }{}
  \end{docCommand}
 
  
 
\begin{docCommand}{if@noparitem}{}
 \end{docCommand}
 \begin{docCommand}{ if@noparitemtrue}{}
 \end{docCommand}
 \begin{docCommand}{ if@noparitemfalse}{}
 \end{docCommand}
 
\begin{teX}
\newif\if@noparitem \@noparitemfalse
\end{teX}
 

\begin{docCommand} {if@noparlist} { }
  \end{docCommand}
  
\begin{docCommand} {if@noparlistfalse}{}
\end{docCommand}

\begin{docCommand} {if@noparlisttrue}{}
\end{docCommand}  
  
\begin{teX}
\newif\if@noparlist \@noparlistfalse
\end{teX}


\begin{docCommand} {if@noitemarg} { }
\end{docCommand}

\begin{docCommand}{if@noitemargtrue}{}
\end{docCommand}  

\begin{docCommand}{if@noitemargfalse}{}
\end{docCommand}  

    \begin{teX}
\newif\if@noitemarg \@noitemargfalse
    \end{teX}


\begin{docCommand}{if@newlist}{}
\end{docCommand}

\begin{docCommand}{if@newlisttrue}{}
\end{docCommand}  

\begin{docCommand}{if@newlistfalse}{}
   \docAuxCommand{@newlist}
\end{docCommand}  

\begin{docCommand}{if@nmbrlist}{}
Checks if the list is to be numbered or not
\end{docCommand}
  
\begin{teX}
\newif\if@nmbrlist  \@nmbrlistfalse
\end{teX}




\begin{docCommand} {@itemlabel} { }
  Temporary definition to store the label.
\end{docCommand}

\section{The \textbackslash list environment}

Notice that the list takes two arguments the first parameter contains instructions as to how
to typeset items, that do not have an |\item[]|. If we need to step a counter we tell it here


\begin{texexample}{The list}{k:list}
\@ifundefined{c@steps}{\newcounter{steps}}{}
\begin{list}{\bfseries\upshape Step \arabic{steps}:}
{%
\usecounter{steps}
\setlength{\labelwidth}{2cm}\setlength{\leftmargin}{2.6cm}
\setlength{\labelsep}{0.5cm}\setlength{\rightmargin}{1cm}
\setlength{\parsep}{0.5ex plus0.2ex minus0.1ex}
\setlength{\itemsep}{0ex plus0.2ex minus0pt}\relax \slshape %
}
\item Understand how they are being build
\item Define your own environment
\item[Has Own] test
\end{list}


\newenvironment{myplain}
{ 
  \list{\bfseries\arabic{steps} }{\usecounter{steps}\bfseries } }
 {\endlist}


\begin{myplain}
\item \lorem

\item[Something]  \lorem
\end{myplain}
\end{texexample}

Looking at the |list| environement differntly it is actually a settings environment, it sets what to do if the item has nor argument and also sets the counter to be used the current levels. It end issuing a parshape command that is going to
shape the paragraph that follows. Next we need to see how item works in more detail. The calculations and some of the work is
delegated to trivlist.

The code follows:


\index{kernel>lists>\textbackslash list}
\begin{docCommand} {list} {\marg{item label } \marg {commands} }
 List takes two arguments and is factory command for building
other lists.
\end{docCommand}


Check that the list is not nested too far.
 
    \begin{teX}
\def\list#1#2{%
  \ifnum \@listdepth >5\relax
    \@toodeep
  \else
    \global\advance\@listdepth\@ne
  \fi
   \end{teX}
Initialize all margins to 0pt in case the user did not specify anything.   
   \begin{texcode}{}{}  
  \rightmargin\z@
  \listparindent\z@
  \itemindent\z@
  \csname @list\romannumeral\the\@listdepth\endcsname
   \end{teX}

Define the item label here, if it is empty it will be defined to empty   
   \begin{texcode}  
   % store the itemlabel 
  \def\@itemlabel{#1}%
  \let\makelabel\@mklab
  \@nmbrlistfalse
  % insert other parameters, spaces, fonts etc
   #2\relax
  % call trivlist to do some calculations
  % and set conditionals 
  \@trivlist
  
  % redefine parskip and parindent
  \parskip\parsep
  \parindent\listparindent
  
  % set the linewidth
  \advance\linewidth -\rightmargin
  \advance\linewidth -\leftmargin
  \advance\@totalleftmargin \leftmargin
  \end{texcode}
  
  Finally set the parshape commands. Thsi is perhaps the most misunderstood line in the whole
  routine. As you see we cannot do any typsetting as yet. Both parameters are just commands to be
  entered before the text, is typeset. Probably you guessed, but what follows is |\item|. At this
  point it will just activate an error if it is missing.  
  \begin{teX}  
  \parshape \@ne \@totalleftmargin \linewidth
  \ignorespaces}
    \end{teX}


\begin{docCommand}{par@deathcycles}{}
\end{docCommand}
    \begin{teX}
\newcount\par@deathcycles
    \end{teX}
 

 \begin{docCommand}{@trivlist}{}
   Auxiliary macro call by trivlist.
 \end{docCommand}
  Because |\par| is sometimes made a no-op it is possible for a missing
 |\item| to produce a loop that does not fill memory and so never gets
 trapped by \TeX.  We thus need to trap this here by seting |\par| to
 count the number of times a paragraph ii is called with no progress
 being made started.
    \begin{teX}
\def\@trivlist{%
  \if@noskipsec \leavevmode \fi
  \@topsepadd \topsep
  \ifvmode
    \advance\@topsepadd \partopsep
  \else
    \unskip \par
  \fi
  \if@inlabel
    \@noparitemtrue
    \@noparlisttrue
  \else
    \if@newlist \@noitemerr \fi
    \@noparlistfalse
    \@topsep \@topsepadd
  \fi
  \advance\@topsep \parskip
  \leftskip \z@skip
  \rightskip \@rightskip
  \parfillskip \@flushglue
  \par@deathcycles \z@
  \@setpar{\if@newlist
             \advance\par@deathcycles \@ne
             \ifnum \par@deathcycles >\@m
               \@noitemerr
               {\@@par}%
             \fi
           \else
             {\@@par}%
           \fi}%
  \global \@newlisttrue
  \@outerparskip \parskip}
    \end{teX}

The function of a |trivlist| is very simple, it sets numbering to false and initializes all  widths to zero, thus at this stage it will default to a normal paragraph then it calls the auxiliary function \refCom{@trivlist}, which will set
the top and bottom skips and other paragraphing parameters, if set.

\begin{texexample}{Trivlist}{ex:trivlist}
\lorem
\begin{trivlist}
\item[] \lorem

\end{trivlist}
\lorem

\end{texexample}

 \begin{docCommand}{trivlist} { \meta{void} }
 \end{docCommand} 
     \begin{teX}
\def\trivlist{%
  \parsep\parskip
  \@nmbrlistfalse
  \@trivlist
  \labelwidth\z@
  \leftmargin\z@
  \itemindent\z@
    \end{teX}

    We initialise |\@itemlabel| so that a \texttt{trivlist} with
    an |\item| not having an optional argument doesn't produce an
    error message.

 \begin{docCommand} {makelabel} {\marg{item label}}
 \end{docCommand}
    \begin{teX}
  \let\@itemlabel\@empty
  \def\makelabel##1{##1}}
    \end{teX}


 \begin{docCommand}{endlist} {  \meta{void}}
  ends the list environment
 \end{docCommand}
     \begin{teX}
\def\endlist{%
  \global\advance\@listdepth\m@ne
  \endtrivlist}
    \end{teX}
 

    The definition of \refCom{trivlist} used to be in ltspace.dtx 
    so that other commands could be `let to it'.  
    They now use \refCom{def}.
 \begin{docCommand}{endtrivlist}{}
 \end{docCommand}
 
% \changes{v1.2b ltspace}{1994/11/12}{Changed order of tests to make
% \refCom{@noitemerror} correct: end of an era.}
% \changes{v1.0i}{1995/05/25}{Macros moved from ltspace.dtx}
% \changes{v1.0n}{1996/10/25}{Change \refCom{indent} to \refCom{leavevmode}}
% \changes{v1.0n}{1996/10/25}{Reset flags explicitly}
% \changes{v1.0o}{1996/10/26}{Correct typo}
    \begin{teX}
\def\endtrivlist{%
  \if@inlabel
    \leavevmode
    \global \@inlabelfalse
  \fi
  \if@newlist
    \@noitemerr
    \global \@newlistfalse
  \fi
  \ifhmode\unskip \par
    \end{teX}
    We also check if we are in math mode and issue an error message
    if so (hoping that |\@currenvir| resolves suitably). Otherwise
    the usual ``perhaps a missing item'' error will get triggered
    later which is confusing.
 \changes{v1.0s}{2002/10/28}{Check for math mode (pr/3437)}
    \begin{teX}
  \else
    \@inmatherr{\end{\@currenvir}}%
  \fi
  \if@noparlist \else
    \ifdim\lastskip >\z@
      \@tempskipa\lastskip \vskip -\lastskip
      \advance\@tempskipa\parskip \advance\@tempskipa -\@outerparskip
      \vskip\@tempskipa
    \fi
    \@endparenv
  \fi
}
    \end{teX}
 

  \begin{docCommand}{@endparenv}{}
 \end{docCommand}
 
     \begin{teX}
\def\@endparenv{%
  \addpenalty\@endparpenalty\addvspace\@topsepadd\@endpetrue}
    \end{teX}


 
\begin{docCommand}{@doendpe}{}
   To suppress the paragraph indentation in text immediately following
   a paragraph-making environment, \cs{everypar} is changed to remove the
   space, and \cs{par} is redefined to restore \cs{everypar}.  Instead of
   redefining \refCom{par} and \cs{everypar}, \refCom{@endparenv} was changed to 
   set the \refCom{@endpe} switch, letting |end| redefine |par| and 
   |everypar|.   This allows paragraph-making environments to work right when called 
 by other environments. )
\end{docCommand}
 

    \begin{teX}
\def\@doendpe{\@endpetrue
     \def\par{\@restorepar\everypar{}
          \par\@endpefalse}\everypar
          {{\setbox\z@\lastbox}\everypar{}\@endpefalse}}
    \end{teX}
    
    Use |\setbox0=\lastbox| instead of   |\hskip -\parindent|   
    so that a |\noindent| becomes a no-op when used before 
    a line immediately following a list environment(23 Oct 86).
 \changes{v1.0k}{1995/11/07}{Enclosed \refCom{setbox0} assignment by a
 group so that it leaves the contents of box $0$ intact.
    }

  \begin{docCommand}{if@endpe}{}
 \end{docCommand}
  \begin{docCommand}{@endpelttrue}{}
 \end{docCommand}
% \begin{macro}{\if@endpe}
% \begin{macro}{\@endpefalse}
% \begin{macro}{\@endpeltrue}
    \begin{teX}
\newif\if@endpe
\@endpefalse
    \end{teX}
% \end{macro}\end{macro}\end{macro}

 \begin{docCommand}{@mklab}{}
 \end{docCommand}
    \begin{teX}
\def\@mklab#1{\hfil #1}
    \end{teX}

 \changes{LaTeX2.09}{1992/09/18}
     {(RmS) Added warning if \refCom{item} is used in math mode}
 \changes{v1.0c}{1994/04/28}
     {Replaced \refCom{@ltxnomath} by \refCom{@inmatherr}}
 \changes{v1.0d}{1994/05/03}
     {Removed superfluous braces}
     
\section{Typesetting the label}     
 
 \begin{docCommand}{item}{\oarg{label}}
 
 \end{docCommand}     
 
 
     \begin{teX}
\def\item{%
  \@inmatherr\item
  \@ifnextchar [\@item{\@noitemargtrue \@item[\@itemlabel]}}
    \end{teX}
 
 \begin{docCommand} {@donoparitem} { }
 \end{docCommand}
    \begin{teX}
\def\@donoparitem{%
  \@noparitemfalse
  \global\setbox\@labels\hbox{\hskip -\leftmargin
                               \unhbox\@labels
                                \hskip \leftmargin}%
  \if@minipage
    \else
      \@tempskipa\lastskip
      \vskip -\lastskip
      \advance\@tempskipa\@outerparskip
      \advance\@tempskipa -\parskip
      \vskip\@tempskipa
  \fi}
    \end{teX}

 
 \begin{docCommand}{@item}{\marg{label} }
 \end{docCommand}
 
     \begin{teX}
\def\@item[#1]{%
  \if@noparitem
    \@donoparitem
  \else
    \if@inlabel
      \indent \par
    \fi
    \ifhmode
      \unskip\unskip \par
    \fi
    \if@newlist
      \if@nobreak
        \@nbitem
      \else
        \addpenalty\@beginparpenalty
        \addvspace\@topsep
        \addvspace{-\parskip}%
      \fi
    \else
      \addpenalty\@itempenalty
      \addvspace\itemsep
    \fi
    \global\@inlabeltrue
  \fi
  \everypar{%
    \@minipagefalse
    \global\@newlistfalse
    \end{teX}
    This |\if@inlabel| check is needed in case an item starts of
    inside a group so that |\everypar| does not become empty
    outside that group. 
 \@nobreakfalse, etc etc.
    \begin{teX}
    \if@inlabel
      \global\@inlabelfalse
    \end{teX}
    The paragraph indent is now removed by using |\setbox...| since
    this makes |\noindent| a no-op here, as it should be. Thus the
    following comment is redundant but is left here for the sake of
    future historians:
    this next command was changed from an hskip to a kern to avoid
    a break point after the parindent box: the skip could cause a
    line-break if a very long label occurs in raggedright setting.
 \changes{v1.0d}{1994/05/03}{\refCom{hskip} changed to \refCom{kern}}
 \changes{v1.0m}{1996/10/23}{\refCom{kern...} changed to \refCom{setbox...}}
 \changes{v1.0r}{1997/02/21}
    {\refCom{ifvoid} check added for \refCom{noindent}. latex/2414}
 If |\noindent| was used after |\item| want to cancel the |\itemindent|
 skip. This case can be detected as the indentation box will be void.
    \begin{teX}
      {\setbox\z@\lastbox
       \ifvoid\z@
         \kern-\itemindent
       \fi}%
    \end{teX}

    \begin{teX}
      \box\@labels
      \penalty\z@
    \fi
    \end{teX}
    This code is intended to prevent a page break after the first
    line of an item that comes immediately after a section title. It
    may be sensible to always forbid a page break after one line of
    an item?  As with all such settings of |\clubpenalty| it is local
    so will have no effect if the item starts in a group.

    Only resetting |\@nobreak| when it is true is now
    essential since now it is sometimes set locally.
 \changes{v1.0m}{1996/10/23}{Added setting of \refCom{clubpenalty} and
    set \refCom{@nobreakfalse} only when necessary}
    \begin{teX}
    \if@nobreak
      \@nobreakfalse
      \clubpenalty \@M
    \else
      \clubpenalty \@clubpenalty
      \everypar{}%
    \fi}%
    \end{teX}
 \changes{v1.0l}{1996/07/26}{Remove unecessary \refCom{global} before
                 \refCom{@nobreak...}}
 \changes{v1.0m}{1996/10/23}{\refCom{@nobreak...} moved into the
          \refCom{everypar} and not executed unconditionally, see above} 
    \begin{teX}
  \if@noitemarg
    \@noitemargfalse
    \if@nmbrlist
    \end{teX}
 \changes{v1.0g}{1995/05/17}{Removed surplus braces}
    \begin{teX}
      \refstepcounter\@listctr
    \fi
  \fi
    \end{teX}
    
Before the label is typeset it is placed in a bo. The latex Team used an |\sbox| so that colour commands can be supported.
It call |\makelabel| with the item parameters.

    \begin{teX}
  \sbox\@tempboxa{\makelabel{#1}}%
  \global\setbox\@labels\hbox{%
    \unhbox\@labels
    \hskip \itemindent
    \hskip -\labelwidth 
    \hskip -\labelsep
    \ifdim \wd\@tempboxa >\labelwidth
       \box\@tempboxa
\end{teX}

\begin{teX}
%
% \changes{LaTeX2.09}{1991/11/22}
%         {(RmS) Changed second call to \refCom{makelabel} to
%           \refCom{unhbox}\refCom{@tempboxa}.
%          Avoids problems with side effects in \refCom{makelabel} and is
%               more efficient.}
%    \begin{teX}
    \else
      \hbox to\labelwidth {\unhbox\@tempboxa}%
    \fix
    \hskip \labelsep}%
  \ignorespaces}
    \end{teX}
 
 
 
 
 \begin{docCommand}{makelabel}{}
 Initial definition of \cs{makelabel} set to produce the infamous and often encountered error, if it
 has not been defined by one of the paragraphing environments.
 \end{docCommand}
 
    \begin{teX}
\def\makelabel#1{%
  \@latex@error{Lonely \string\item--perhaps a missing
        list environment}\@ehc}
    \end{teX}
 

 \begin{docCommand}{nbitem}{}
 \end{docCommand}
% \begin{macro}{\@nbitem}
 \changes{v1.0g}{1995/05/17}{Removed surplus braces}
    \begin{teX}
\def\@nbitem{%
  \@tempskipa\@outerparskip
  \advance\@tempskipa -\parskip
  \addvspace\@tempskipa}
    \end{teX}

 \begin{docCommand}{usecounter}{}
 \end{docCommand}
    \begin{teX}
\def\usecounter#1{\@nmbrlisttrue\def\@listctr{#1}\setcounter{#1}\z@}
    \end{teX}


 \section{Enumerate}

  Enumeration is done with four counters: |enumi|, |enumii|, |enumiii|
  and |enumiv|, where |enum|N controls the numbering of the Nth level
  enumeration.  The label is generated by the commands
  \refCom{labelenumi} \ldots{} \cs{labelenumiv}, which should be defined
  by the document style.
  
  Note that \cs{p@enum}N \cs{theenum}N defines the output
  of a \refCom{ref} command.  A typical definition might be:
  
 \begin{verbatim}
     \def\theenumii{\alph{enumii}}
     \def\p@enumii{\theenumi}
     \def\labelenumii{(\theenumii)}
 \end{verbatim}
 which will print the labels as `(a)', `(b)', \ldots
 and print a \refCom{ref} as `3a'.

 The item numbers are moved to the right of the label box, so they are
 always a distance of \refCom{labelsep} from the item.

 \refCom{@enumdepth} holds the current enumeration nesting depth.

 Itemization is controlled by four commands: \docAuxCommand{labelitemi},
 \docAuxCommand{labelitemii},
 \docAuxCommand{labelitemiii}, and \docAuxCommand{labelitemiv}.
 To cause the second-level list to be
 bulleted, you just define \docAuxCommand{labelitemii}
 to be |$\bullet$|.  |@itemspacing|
 and \refCom{@itemdepth} are the analogs of |@enumspacing| and
 \refCom{@enumdepth}. These are not defined in the kernel, their definition is delegated to the classes and to another extend to the users.


 \begin{docCommand}{@enumdepth}{}
  The enumeration depth. 
 \end{docCommand}
     \begin{teX}
\newcount\@enumdepth \@enumdepth = 0
    \end{teX}
 

Next the counters for the \refEnv{enumerate} environment, \docCounter{c@enumi}, \docCounter{c@enumii}, \docCounter{c@enumiii} and \docCounter{c@enumiv} are created.

    \begin{teX}
 \@definecounter{enumi} 
 \@definecounter{enumii} 
 \@definecounter{enumiii} 
 \@definecounter{enumiv}
    \end{teX}


 \begin{docEnvironment}{enumerate}{}{}
 \end{docEnvironment}  
   The enumerate environment enumerates the list. The macro
  is written very efficiently and defines the basic structure. The
  typesetting parameters are left for the class files such as book
  to define them.

     \begin{teX}
 \def\enumerate{%
  \ifnum \@enumdepth >\thr@@\@toodeep\else
    \advance\@enumdepth\@ne
    \edef\@enumctr{enum\romannumeral\the\@enumdepth}%
      \expandafter
      \list
         \csname label\@enumctr\endcsname
        {\usecounter\@enumctr\def\makelabel##1{\hss\llap{##1}}}%
  \fi}
    \end{teX}

    \begin{teX}
 \let\endenumerate =\endlist
    \end{teX}

 \begin{docCommand}{@itemdepth}{}
 \end{docCommand}
    \begin{teX}
\newcount\@itemdepth \@itemdepth = 0
    \end{teX}

\section{The itemize environment}

Finally the \docValue{itemize} environment is defined. This is defined in the kernel to have a limit of 4 levels only.

Itemization is controlled by four commands, which are defined in the standard classes: |\labelitemi|,
   |\labelitemii|, |\labelitemiii|, and |\labelitemiv|, which define
    the labels of the various itemization levels: the symbols used in the standard classes are
    bullet, bold en-dash, centered asterisk and centred dot.



\begin{docEnvironment}{itemize}{}
\end{docEnvironment}


     \begin{teX}
 \def\itemize{%
  \ifnum \@itemdepth >\thr@@\@toodeep\else
    \advance\@itemdepth\@ne
    \edef\@itemitem{labelitem\romannumeral\the\@itemdepth}% (*@\dcircle{1}@*)
    \end{teX}

The \dcircle{1} is a hack to create the |\@itemi|\ldots|\@itemv| names. How it works is that the
|\romannumeral| will expand until it finds a number and leave the letter |i| etc. in the stream.

\begin{texexample}{Creating a suffix with Roman Numerals}{ex:rmnumeral}
\newcounter{tempcounter}
\setcounter{tempcounter}{5}
\edef\itemtest{labelitem\romannumeral\thetempcounter}
%  the replacement
\meaning\itemtest

\itemtest
\end{texexample}
    
    
A romannumeral just returns nothing if it encounters the number \texttt{0} or a negative number. 

\begin{texexample}{Creating a suffix with Roman Numerals}{ex:rmnumeral1}
% counter has been created globally
% commented out.
%\newcounter{tempcounter}
\setcounter{tempcounter}{0}
\edef\itemtest{labelitem\romannumeral\thetempcounter}
%  the replacement
\meaning\itemtest\par
\itemtest
\end{texexample}
    
    
Call the \refCom{list} sets all parameters and defines a  \refCom{makelabel} for this particular
environment. This is a good place to also hook other items to decorate the label, if necessary.
Remember also that the \docAuxCommand{@itemitem}, defined above is just there to facilitate  
calling |labelitemi| \ldots |labelitemn| depending on the level.

\begin{teX}[numbers=none]
    \expandafter
    \list
      
      \csname\@itemitem\endcsname
      {\def\makelabel##1{\hss\llap{##1}}}%
  \fi}
\end{teX}

The astude reader will notice at this point that although we defined the name of labeli etc, they still have no meaning. As mentioned earlier there is still some work to be done before we have full lists. These are defined in the standatrd classes and the definitions are shown below.\tcbdocmarginnote{revised\\13-06-2018}

\begin{teX}
\newcommand\labelitemi{\textbullet}
\newcommand\labelitemii{\normalfont\bfseries \textendash}
\newcommand\labelitemiii{\textasteriskcentered}
\newcommand\labelitemiv{\textperiodcentered}
\end{teX}

Finally the end of the environment is defined.

    \begin{teX}
\let\enditemize =\endlist
    \end{teX}



