\chapter{Colors}
\section{Introduction}

One area where \latex shows its age is the manipulation and use of colors. When \tex was developed color printing was not possible and the introduction of colors is now only predominantly via the use of the |xcolor| package. Despite such limitations colors can be used fairly easily.

\newthought{The figure below, shows the wavelengths} in nm of the visible light. It has been drawn using the \docpkg{xcolor} package and the native \latex environment \cs{picture}. The colors can be typest using the wavelength of light.

\smallskip

\begin{texexample}{}{}
  \hbox{\color{thered} A TesT}
\end{texexample}




\newcount\WL \unitlength.75pt

\begin{figure}
\hskip-3pt\scalebox{0.9}{
\noindent

\begin{picture}(460,60)(355,-10)
\sffamily \tiny \linethickness{1.25\unitlength} \WL=360
\multiput(360,0)(1,0){456}%
{{\color[wave]{\the\WL}\line(0,1){50}}\global\advance\WL1}
\linethickness{0.25\unitlength}\WL=360
\multiput(360,0)(20,0){23}%
{\picture(0,0)
\line(0,-1){5} \multiput(5,0)(5,0){3}{\line(0,-1){2.5}}
\put(0,-10){\makebox(0,0){\the\WL}}\global\advance\WL20
\endpicture}
\end{picture}}
\caption{The visible spectrum nm}
\end{figure}

The |xcolor| package provides numerous macros for typesetting colors, using a variety of methods and color schemes. For example we can use the command \cs{color} to print a text sample in color.
\newlength\pull

\def\colorSample#1{%
\leavevmode
\parindent0pt
   \def\colorRule{\color[wave]{#1}\rule{\textwidth}{0.4pt}} 
   \colorRule
%% set to the width of the box
   \settowidth\pull{\framebox{\Large #1 nm}}
%% pull by one em
   \addtolength\pull{1em}
   \hskip -\pull{\color[wave]{#1}{{\framebox{\Large #1 nm}}}}%
%% add story
   \hskip1em\noindent\onepar\par
   \colorRule
}
\bgroup
\colorSample{385}
\colorSample{809}
\egroup


\section{Specifying colors by name}

The easier way to specify colors is to use the pre-build names available
with the package drivers.


\section{Color boxes}

Now and then users want to place text in colorboxes. The macro \cs{colorbox} can be used to provide background text to a box.

\begin{macro}{\colorbox}
The colorbox macro takes one command. As it is modelled on the same concept as an
\cs{fbox} it will add a bit of space around the containing text. If you want
to remove it you will have to set the \cs{fboxsep=0pt}.
\end{macro}


\begin{texexample}{Color boxes}{}
\fboxsep0pt
\colorbox{green}{\begin{minipage}{3cm}
  Some text
\end{minipage}}

\fboxrule.2pt\fboxsep-.2pt
\fbox{\begin{minipage}{3cm}
  Some text
\end{minipage}}
\end{texexample}

\begin{macro}{\textcolor}
The \cs{textcolor} takes two arguments and typesets the contents of the second argument, using the color specified in the first argument.
\end{macro}

\begin{texexample}{Color text}{}
\textcolor{blue}{This is typeset in blue color.}

\end{texexample}

\section{Mixing colors}

Colors can be mixed by using a convention using (!).

\begin{texexample}{Mixing colors}{}
\color{blue!40!yellow}
\lorem
\end{texexample}

One can also mix colors from different color schemes and |xcolor| will take care of the calculations.

\section{Defining colors}

There are two commands that one can use to define colors.

\begin{macro}{\definecolor}\oarg{type}\marg{name}\marg{model-list}\marg{spec-list} 
The |\definecolor| command is one of the commands that may be used to assign a \textit{name} to a specific color. Afterwards, this color is known to the system (in the current group) and may be used in color expressions. Note that an existing color name will be overwritten.
\end{macro} 

\begin{texexample}{Defining colors}{}
\definecolor{red}{rgb}{1,0,0}
\definecolor{mygray}{named}{black}
\colorlet{myblack}{black}

\color{myblack}\lorem
\end{texexample}










