
%%% DOCUMENATION MACROS
%% note becareful with styles here

\thispagestyle{plain}
\cxset{image=breakerboys, custom=stewart}
\chapter{Documentation}


\section{Documentation macros}

When developing this package the need arose to define a number of documentation macros. I have used heavily macros and ideas present in the doc package, pgf documentation, biblatex documentation  and tcolorbox and for which I am grateful to their respective authors. The major change was to adopt the macros to use different fonts and colors and to use these from a list of key values defined at document level. More about this later. General package user documentation as opposed to package documentation that can be achieved using the doc/docstrip system requires that macros and environments be developed for the following:

\begin{enumerate}
\item Macros for command documentation.
\item Environments for commands and options.
\item Latex examples that need to be executed within the document as well as described.
\end{enumerate}

\subsection{Color management}
One of the first requirements for redefining some of the standard doc commands is the need to use color easily, hence we will try and define a certain amount of keys for colors.

Just a bit of a refresher, to define colors we use, either the \cs{definecolor} or the \cs{colorlet} commands.

\emphasis{definecolor,colorlet}
\begin{tcolorbox}
\begin{lstlisting}
% used for hyperlinks
\definecolor{Hyperlink}{rgb}{0.281,0.275,0.485}
\colorlet{thehyperlink}{theblue}
\end{lstlisting}
\end{tcolorbox}

We use a semantic approach, where the colors are first defined with a mnemonic command such as {\bfseries\textcolor{theblue}{theblue}} and then we define a semantic command such as the\cs{option} that lets the color to the option command. This sort of double entry has proved useful in navigating through the dozen of the commands that I needed for this documentation.

%\definecolor{theblue} {rgb}{0.02,0.04,0.48}
%\definecolor{thered}  {rgb}{0.65,0.04,0.07}
%\definecolor{thegreen}{rgb}{0.06,0.44,0.08}
%\definecolor{thelightgreen}{rgb}{0.06,0.44,0.06}
%\definecolor{thegrey} {gray}{0.5}
%\definecolor{thegrey} {gray}{0.5}
%\definecolor{theshade}{gray}{0.94}
%\definecolor{theframe}{gray}{0.75}
%\definecolor{thecream}{rgb}{1,0.95,0.4}
%\definecolor{spot}{rgb}{0,0.2,0.6}
%\definecolor{boxframe}{gray}{0.8}
%\definecolor{boxfill}{rgb}{0.95,0.95,0.99}
%\definecolor{theoption}{rgb}{0.118,0.546,0.222}
%\definecolor{themacro}{rgb}{0.784,0.06,0.176}
%\definecolor{ExampleFrame}{rgb}{0.628,0.705,0.942}
%\definecolor{ExampleBack}{rgb}{0.963,0.971,0.994}
%\definecolor{Hyperlink}{rgb}{0.281,0.275,0.485}
%\colorlet{thehyperlink}{theblue}
%\newcommand*{\defaultcolor}{\color{black}}
%\newcommand*{\spotcolor}{\color{spot}}

\subsection{Semantic color names}
\begin{marglist}
\item [\option{theoption}] Coloring of options in margin lists.
\item [\option{themacro}] Coloring of command macros \cs{foo}.
\item [\option{hyperlink}] If we use the \texttt{hyperref} package a number of colors need to be defined for links.
\end{marglist}

\subsection{Named colors}
Standard colors that we provide are:
\begin{marglist}
\item [\textcolor{theblue}{theblue}] This color is used mainly for options.
\item [\textcolor{thered}{thered}] The color mostly used for macro commands and keys.
\item [\textcolor{thegreen}{thegreen}] used for environments.
\item [\textcolor{thelightgreen}{thelightgreen}] Used for margin lists.
\item [\textcolor{thegray}{thegray}] Used as a background to the listings.
\item [\colorbox{thegrey}{\color{white}thegrey}] Alias for the gray to satisfy both sides of the Atlantic and as I sometimes don't remeber which is which.
\item [\colorbox{theshade}{theshade}] Another slightly lighter shade.
\end{marglist}



\begin{marglist}
\item [\cs{cs}] \cs{cs}\marg{text} Prints a command.
\item [\cs{cmd}] \cmd{\cmd}\marg{\cmd{\foo}} Prints a command.
\item[\cs{meta}] \marg{macro|text} Prints an argument
\end{marglist}




\section{Lists for documentation}



The environment \env{marglist}
\begin{marglist}
\item[testing]\lorem
\item [test]\lorem
\end{marglist}

\env{keymarginlist}This environment is suitable for listing keys, set-in the margin.

\begin{keymarglist}
\item[bibliography] The term <bibliography>, also available as \cmd{bibname}.
\item[references] The term <references>, also available as \cmd{refname}.
\item[shorthands] The term <list of shorthands> or <list of abbreviations>, also available as \cmd{losname}.
\end{keymarglist}


\env{argumentlist} This environment is suitable for listing macro arguments and their explanations.

\newenvironment*{argumentlist}[1]
  {\list{}{%
     \settowidth{\labelwidth}{\displayverbfont#1}%
     \setlength{\labelsep}{1em}%
     \setlength{\leftmargin}{\labelwidth}%
     \addtolength{\leftmargin}{\labelsep}%
     \setlength{\itemsep}{0pt}%
     \renewcommand*{\makelabel}[1]{\displayverbfont##1\hss}}}
  {\endlist}

\begin{argumentlist}{00}
\item[\#1] The last names. If a name consists of a single part only (for example, <Aristotle>), this part will be treated as the last name.
\item[\#2] The last names, given as initials.
\item[\#3] The first names. This argument also includes all middle names.
\end{argumentlist}

\begin{tabular}{ll}
\env{marglist} & description list\\
\end{tabular}


\section{Future improvements}

Future improvements will involve defining additional environments and styles to integrate styles to those provided for chapters. Treat the code so far as a proof of concept experiment. It is also possible to improve on the code by utilizing some of the common packages.

