\chapter{Installation}


\section{Using the latest version}

fixltx2e, etex, morefloats, latexrelease are all affected to some extent.

At each LaTeX release there is a one or two page newsletter describing the main changes (ltnews 22 and 23 for the 2015/01/01 and 2015/10/01 releases, respectively) these days we bundle them as one file texdoc ltnews or ltnews.pdf from CTAN.

As detailed in ltnews 23 fixltx2e is obsolete, all the fixes that were in that package have now been applied to the format and instead a new package, latexrelease, allows you to back them out if needed.

Also in the January release, LaTeX started using extended register allocations by default and introduced a new \extrafloats command to access more float boxes (for figures and tables, etc.) so etex.sty shouldn't normally be used and the contributed morefloats package isn't really needed (although it's been updated to use \extrafloats internally if defined so using it does no harm).

As noted in ltnews 23, the October release incorporated a lot of support for LuaTeX, mostly a refactoring of the existing contributed luatexbase code so luatexbase should not normally be used. (luatexbase as distributed is now a stub that defines a thin compatibility layer for some of the older commands that were not exactly copied, but otherwise uses the new code.)

So of these only really etex.sty shouldn't be used. If it is used, then in order to get maximum compatibility it overwrites the new register allocation mechanism with its original one, so losing many of the new features.

fixltx2e is harmless if used, it does nothing and just gives a warning message.

morefloats is fine, it uses the new mechanisms internally if used on a new engine, it's just that the problems it solved are less likely to occur now, and the \extrafloats syntax can be used.

luatexbase is also fine, it adds a compatibility layer emulating the older package of the same name but does not affect code written to the new LaTeX base LuaTeX support.