
\chapter{lterror.dtx}
\label{kernel:lterror}

 \section{Error handling}

 This section defines \LaTeXe's error commands. Most of the error
 messages are defined here, although now and then some error
 messages are defined in other |.dtx| files.

 The `2ekernel' code ensures that a |\usepackage{autoerr}| is
 essentially ignored if a `full' format is being used that has
 the error messages already in the format.
    \begin{teX}
    \expandafter\let\csname ver@autoerr.sty\endcsname\fmtversion
    \end{teX}


 \subsection{General commands}

 \begin{docCommand}{MessageBreak}{}
    This command prints a new-line inside a message, followed by a 
    continuation line beginning with |\@msg@continuation|.  Normally it is 
    defined to be |\relax|, but inside messages, it is let to 
    \refCom{@message@break} for example in the definition of \refCom{GenericWarning}.
 \end{docCommand}
    \begin{teX}
\let\MessageBreak\relax
    \end{teX}
  

 \begin{docCommand}{GenericInfo}{}
    This takes two arguments: a continuation and a message, and sends 
    the result to the log file.
    \begin{teX}
\DeclareRobustCommand{\GenericInfo}[2]{%
   \begingroup
      \def\MessageBreak{^^J#1}%
      \set@display@protect
      \immediate\write\m@ne{#2\on@line.}%
   \endgroup
}
    \end{teX}
 \end{docCommand}



 \begin{docCommand}{GenericWarning}{}
    This takes two arguments: a continuation and a message, and sends 
    the result to the screen.
    \begin{teX}
\DeclareRobustCommand{\GenericWarning}[2]{%
   \begingroup
      \def\MessageBreak{^^J#1}%
      \set@display@protect
      \immediate\write\@unused{^^J#2\on@line.^^J}%
   \endgroup
}
    \end{teX}
  \end{docCommand}

 \begin{docCommand}{GenericError}{}
 This macro takes four arguments: a continuation,
 an error message, where to go for further information, and the help
 information.  It displays the error message, and sets the error help
 (the result of typing |h| to the prompt), and does a horrible hack
 to turn the last context line (which by default is the only context
  line) into just three dots.  This could be made more efficient.

    \begin{teX}
\def\GenericError{\@autoerr\GenericError}

\bgroup
\lccode`\@=`\ %
\lccode`\~=`\ %
\lccode`\}=`\ %
\lccode`\{=`\ %
\lccode`\T=`\T%
\lccode`\H=`\H%
\catcode`\ =11\relax%
\lowercase{%
\egroup%
    \end{teX}

 Unfortunately \TeX\ versions older than 3.141 have a bug which means
 that |^^J| does not force a linebreak in |\message| and |\errmessage|
 commands. So for these old \TeX's we use |\typeout| to produce the
 message, and then have an empty |\errmessage| command. This causes an
 extra line of the form
\begin{verbatim}
 ! .
\end{verbatim}
 To appear on the terminal, but if you do not like it, you can always
 upgrade your \TeX! In order for your format to use this version, you
 must define the macro |\@TeXversion| to be the version number, e.g.,
 3.14 of the underlying \TeX. See the comments in
 \texttt{ltdircheck.dtx}.
    \begin{teX}
\dimen@\ifx\@TeXversion\@undefined4\else\@TeXversion\fi\p@%
\ifdim\dimen@>3.14\p@%
    \end{teX}

 First the `standard case'.
    \begin{teX}
\DeclareRobustCommand{\GenericError}[4]{%
\begingroup%
\immediate\write\@unused{}%
\def\MessageBreak{^^J}%
\set@display@protect%
\edef%
%    %<-------------------do not delete this space!------------------->%
\@err@                                                                 %
{{#4}}%
\errhelp
%    %<-------------------do not delete this space!------------------->%
\@err@                                                                 %
\let
%    %<-------------------do not delete this space!------------------->%
\@err@                                                                 %
\@empty
\def\MessageBreak{^^J#1}%
\def~{\errmessage{%
#2.^^J^^J%
#3^^J%
Type  H <return>  for immediate help%
%    %<-------------------do not delete this space!------------------->%
\@err@                                                                 %
}}%
~%
\endgroup}%
    \end{teX}

    \begin{teX}
\else%
    \end{teX}

 Secondly the version for old \TeX's.
    \begin{teX}
\DeclareRobustCommand{\GenericError}[4]{%
\begingroup%
\immediate\write\@unused{}%
\def\MessageBreak{^^J}%
\set@display@protect%
\edef%
    %<-------------------do not delete this space!------------------->%
\@err@                                                                 %
{{#4}}%
\errhelp
    %<-------------------do not delete this space!------------------->%
\@err@                                                                 %
\let
    %<-------------------do not delete this space!------------------->%
\@err@                                                                 %
\errmessage
\def\MessageBreak{^^J#1}%
\def~{\typeout{! %
#2.^^J^^J%
#3^^J%
Type  H <return>  for immediate help.}%
    %<-------------------do not delete this space!------------------->%
\@err@                                                                 %
{}}%
~%
\endgroup}%
    \end{teX}

    \begin{teX}
\fi}%

    \end{teX}
 \end{docCommand}

\subsection{Package and Class error messages}
 
 
 
 
 \begin{docCommand}{PackageWarningNoLine}{}
 \end{docCommand}
 
 \begin{docCommand}{PackageInfo}{}
 \end{docCommand}
 
 \begin{docCommand}{ClassError}{}
  \end{docCommand}

 \begin{docCommand}{ClassWarning}{}
 \begin{docCommand}{ClassWarningNoLine}{}
 \begin{docCommand}{ClassInfo}{}
  These commands are intended for use by package and class writers, to
  give information to authors.  The syntax is:
    \begin{quote}
       |\PackageError{|\meta{package}|}{|^^A
          \meta{error}|}{|\meta{help}|}| \\
       |\PackageWarning{|\meta{package}|}{|\meta{warning}|}| \\
       |\PackageWarningNoLine{|\meta{package}|}{|\meta{warning}|}| \\
       |\PackageInfo{|\meta{package}|}{|\meta{info}|}|
    \end{quote}
    and similarly for classes.  The |Error| commands print the
    \meta{error} message, and present the interactive prompt; if the
    author types |h|, then the \meta{help} information is displayed.
    The |Warning| commands produce a warning but do not present the
    interactive prompt.  The |WarningNoLine| commands do the same,
    but don't print the input line number.  The |Info| commands write
    the message to the
    |log| file.  Within the messages, the command 
    |\MessageBreak| can be used to
    break a line, |\protect| can be used to protect command names,
    and |\space| is a space, for example:
 \begin{verbatim}
    \newcommand{\foo}{FOO}
    \PackageWarning{ethel}{%
       Your hovercraft is full of eels,\MessageBreak
       and \protect\foo\space is \foo}
 \end{verbatim}
    produces:
 \begin{verbatim}
    Package ethel warning: Your hovercraft is full of eels,
    (ethel)                and \foo is FOO on input line 54.
 \end{verbatim}
 \end{docCommand}
 \end{docCommand}
 \end{docCommand}
 
 The definition of the commands are fairly straight forward.  They all use the \refCmd{GenericError} command to display the error on the terminal. The new
 expl3 module for \latex3e, has expanded on this work tremendously, so that
 there is no real need for using these macros any longer.
  
\begin{docCommand}{PackageError}{ \marg{package} \marg{error} \marg{help}}
 \end{docCommand}
    \begin{teX}
\gdef\PackageError#1#2#3{%
   \GenericError{%
      (#1)\@spaces\@spaces\@spaces\@spaces
   }{%
      Package #1 Error: #2%
   }{%
      See the #1 package documentation for explanation.%
   }{#3}%
}
    \end{teX}
    
    
\begin{docCommand}{PackageWarning}{ \meta{package} \meta{warning}}
 \end{docCommand}
    \begin{teX}
\def\PackageWarning#1#2{%
   \GenericWarning{%
      (#1)\@spaces\@spaces\@spaces\@spaces
   }{%
      Package #1 Warning: #2%
   }%
}
\def\PackageWarningNoLine#1#2{%
   \PackageWarning{#1}{#2\@gobble}%
}
\def\PackageInfo#1#2{%
   \GenericInfo{%
      (#1) \@spaces\@spaces\@spaces
   }{%
      Package #1 Info: #2%
   }%
}
    \end{teX}

    \begin{teX}

\gdef\ClassError#1#2#3{%
   \GenericError{%
      (#1) \space\@spaces\@spaces\@spaces
   }{%
      Class #1 Error: #2%
   }{%
      See the #1 class documentation for explanation.%
   }{#3}%
}
    \end{teX}

    \begin{teX}
\def\ClassWarning#1#2{%
   \GenericWarning{%
      (#1) \space\@spaces\@spaces\@spaces
   }{%
      Class #1 Warning: #2%
   }%
}
\def\ClassWarningNoLine#1#2{%
   \ClassWarning{#1}{#2\@gobble}%
}
\def\ClassInfo#1#2{%
   \GenericInfo{%
      (#1) \space\space\@spaces\@spaces
   }{%
      Class #1 Info: #2%
   }%
}
    \end{teX}



Some errors are marked to be generated only in the \latex2e kernel. This
is where we see the reference to the LaTeX Manual or the LaTeX Companion
for help is mentioned.
 \begin{docCommand}{@latex@error}{}
 \begin{docCommand}{@latex@warning}{}
 \begin{docCommand}{@latex@warning@no@line}{}
 \begin{docCommand}{@latex@info}{}
 \begin{docCommand}{@latex@info@no@line}{}
     \begin{teX}
\gdef\@latex@error#1#2{%
   \GenericError{%
      \space\space\space\@spaces\@spaces\@spaces
   }{%
      LaTeX Error: #1%
   }{%
      See the LaTeX manual or LaTeX Companion for explanation.%
   }{#2}%
}
    \end{teX}

    \begin{teX}
\def\@latex@warning#1{%
   \GenericWarning{%
      \space\space\space\@spaces\@spaces\@spaces
   }{%
      LaTeX Warning: #1%
   }%
}
    \end{teX}

    \begin{teX}
\def\@latex@warning@no@line#1{%
   \@latex@warning{#1\@gobble}}
    \end{teX}

    \begin{teX}
\def\@latex@info#1{%
   \GenericInfo{%
      \@spaces\@spaces\@spaces
   }{%
      LaTeX Info: #1%
   }%
}
    \end{teX}

    \begin{teX}
\def\@latex@info@no@line#1{%
  \@latex@info{#1\@gobble}}
    \end{teX}

    |\@font@warning| and |\@font@info| are defined later since they
    have to be redefined by the \texttt{tracefnt} package.
\begin{verbatim}
\def\@font@warning#1{%
   \GenericWarning{%
      {(font)\@spaces\@spaces}%
      {Font Warning: #1}%
 }
\def\@font@info#1{%
   \GenericInfo{%
      (font)\space\@spaces
   }{%
      Font Info: #1%
   }%
 }
\end{verbatim}
 \end{docCommand}
 \end{docCommand}
 \end{docCommand}
 \end{docCommand}
 \end{docCommand}

 \begin{docCommand}{c@errorcontextlines}{}
 \changes{LaTeX2e}{1993/11/22}{Macro added}
  |\errorcontextlines| as a \LaTeX\ counter, so that it may be be
  manipulated with |\setcounter| (once it is defined :-)
    \begin{teX}
\let\c@errorcontextlines\errorcontextlines
\c@errorcontextlines=-1
    \end{teX}
 \end{docCommand}

 \changes{v1.0d}{1994/03/28}
     {Remove test for \cs{inputlineno} undefined.}
 \begin{docCommand}{on@line}{}
    The message ` on input line~$n$', if possible.
    \begin{teX}
\ifnum\inputlineno=\m@ne
  \let\on@line\@empty
\else
  \def\on@line{ on input line \the\inputlineno}
\fi
    \end{teX}
 \end{docCommand}

  \begin{docCommand}{@warning}{}
  \begin{docCommand}{@@warning}{}
  \begin{docCommand}{@latexerr}{}
     Older \LaTeX{} messages.  For the moment, these
     |\let| to the new message commands.  They may be changed later,
     once only obsolete packages and classes contain them.
 \changes{v1.0b}{1993/12/03}{Set \cs{c@errorcontextlines} to -1}
 \changes{v1.0e}{1993/04/09}{Mention The Companion}
 \changes{v1.0f}{1993/04/11}{Remove setting of errorcontextlines}
 \changes{v1.0k}{1994/05/01}{(CAR) Added draft \cs{@latexinfo}.}
 \changes{v1.0n}{1994/05/10}{(ASAJ) Added extra blank lines to
           \cs{@latexerr}.}
 \changes{v1.0o}{1994/05/11}
     {(ASAJ) Removed one of the extra blank lines to \cs{@latexerr}.}
    \begin{teX}
\let\@warning\@latex@warning
\let\@@warning\@latex@warning@no@line
</2ekernel|autoload>
\global\let\@latexerr\@latex@error
    \end{teX}
  \end{docCommand}
  \end{docCommand}
  \end{docCommand}

 \begin{docCommand}{@spaces}{}
    Four spaces.
    \begin{teX}
\def\@spaces{\space\space\space\space}
    \end{teX}
 \end{docCommand}

 \subsection{Specific errors}

 The four commands
\docAuxCmd{@eha},
\docAuxCmd{@ehb},
\docAuxCmd{@ehc}, 
and \docAuxCmd{@ehd}, 
 expand to specific errors.   The more common error help messages are defined by macros named as a series of |\@eh|\meta{letter}. Obviously short for errorhelp
    \begin{teX}
\gdef\@eha{%
  Your command was ignored.\MessageBreak
  Type \space I <command> <return> \space to replace it %
  with another command,\MessageBreak
  or \space <return> \space to continue without it.}
\gdef\@ehb{%
  You've lost some text. \space \@ehc}
\gdef\@ehc{%
  Try typing \space <return> %
  \space to proceed.\MessageBreak
  If that doesn't work, type \space X <return> \space to quit.}
\gdef\@ehd{%
  You're in trouble here.  \space\@ehc}
    \end{teX}
 As |\latex@error| triggers the autoload, these definitions
 should not be needed in the autoload format, but just to be safe\ldots
    \begin{teX}
%<*autoload>
\let\@eha\@empty\let\@ehb\@empty\let\@ehc\@empty\let\@ehd\@empty
%</autoload>
    \end{teX}


 Here are most of the error message-generating commands of \LaTeX.
 \begin{docCommand}{@autoerr}{}
 Make this autoload command robust, as it may be read in at
 unpredicatble times.
    \begin{teX}
<autoload>\def\@autoerr{\protect\@autoload{err}\protect}
    \end{teX}
 \end{docCommand}

 \begin{docCommand}{@notdefinable}{}
    Error message generated in |\@ifdefinable| from calls
    to one of the commands |\newcommand|, |\newlength| or |\newtheorem|
    specifying an already-defined command name or one that begins
    |\end...|.
 \changes{v1.2n}{1998/05/28}{Added message re `end...' pr/1555}
    \begin{teX}
\gdef\@notdefinable{%
    \end{teX}
 \end{docCommand}

 \begin{docCommand}{@nolnerr}{}
 Generated by |\newline| and |\\| when called in vertical mode.
    \begin{teX}
\gdef\@nolnerr{%
<!autoload>  \@latex@error{There's no line here to end}\@eha}
<autoload>  \@autoerr\@nolnerr}
    \end{teX}
 \end{docCommand}

 \begin{docCommand}{@nocounterr}{}
  Generated by |\setcounter|, |\addtocounter| or
  |\newcounter| if applied to an undefined counter \meta{cnt}.

 \begin{docCommand}{@nocnterr}{}
 Obsolete error message generated in \LaTeX2.09 by
 |\setcounter|, |\addtocounter| or |\newcounter|
 for undefined counter.
 DO NOT use for \LaTeXe\ it MIGHT vanish!
 Use |\@nocounterr|\marg{cnt} instead.
    \begin{teX}
\gdef\@nocounterr#1{%
<!autoload>  \@latex@error{No counter '#1' defined}\@eha}
<autoload>  \@autoerr\@nocounterr}
\gdef\@nocnterr{\@nocounterr?}
    \end{teX}
 \end{docCommand}
 \end{docCommand}

 \begin{docCommand}{@ctrerr}{}
 Called when trying to print the value of a counter
 numbered by letters that's greater than 26.
    \begin{teX}
\gdef\@ctrerr{%
<!autoload>  \@latex@error{Counter too large}\@ehb}
<autoload>  \@autoerr\@ctrerr}
    \end{teX}
 \end{docCommand}


 \begin{docCommand}{@nodocument}
 Error produced if paragraphs are typeset in the preamble.
 \changes{v1.2m}{1996/11/04}{Always define \cs{@nodocument}
           in kernel, so that it can be cleared by \cs{document}.}
    \begin{teX}
<!def>\gdef\@nodocument{%
<!def>  \@latex@error{Missing \protect\begin{document}}\@ehd}
    \end{teX}
 \end{docCommand}

 \begin{docCommand}{@badend}{}
 Called by |\end| that doesn't match its |\begin|.
 RmS 1992/08/24: added code to |\@badend| to display position of
               non-matching |\begin|.
 FMi 1993/01/14: missing space added.
    \begin{teX}
\gdef\@badend#1{%
<!autoload>  \@latex@error{\protect\begin{\@currenvir}\@currenvline
<!autoload>                     \space ended by \protect\end{#1}}\@eha}
<autoload>  \@autoerr\@badend}
    \end{teX}
 \end{docCommand}

 \begin{docCommand}{@badmath}{}
 Called by |\[|, |\]|, |\(| or |\)| when used in wrong mode.
    \begin{teX}
\gdef\@badmath{%
<!autoload>  \@latex@error{Bad math environment delimiter}\@eha}
<autoload>  \@autoerr\@badmath}
    \end{teX}
 \end{docCommand}

 \begin{docCommand}{@toodeep}{}
 Called by a list environment nested more than six levels
 deep, or an enumerate or itemize nested more than four levels.
    \begin{teX}
\gdef\@toodeep{%
<!autoload>  \@latex@error{Too deeply nested}\@ehd}
<autoload>  \@autoerr\@toodeep}
    \end{teX}
 \end{docCommand}

 \begin{docCommand}{@badpoptabs}{}
 Called by |\endtabbing| when not enough |\poptabs| have
 occurred, or by |\poptabs| when too many have occurred.
    \begin{teX}
\gdef\@badpoptabs{%
<!autoload>  \@latex@error{\protect\pushtabs\space and \protect\poptabs
<!autoload>      \space don't match}\@ehd}
<autoload>  \@autoerr\@badpoptabs}
    \end{teX}
 \end{docCommand}

 \begin{docCommand}{@badtab}{}
 Called by |\>|, |\+| , |\-| or |\<| when stepping to an undefined tab.
    \begin{teX}
\gdef\@badtab{%
<!autoload> \@latex@error{Undefined tab position}\@ehd}
<autoload>  \@autoerr\@badtab}
    \end{teX}
 \end{docCommand}

 \begin{docCommand}{@preamerr}{}
 \changes{v1.2k}{1995/10/24}
         {Modify autoload support}
    This error is special: it appears in places where we normally have
    to |\protect| expansions. However, to prevent a protection of
    the error message itself (which would result in the message
    getting printed not issued on the terminal) we need to locally
    reset |\protect| to |\relax|.
    \begin{teX}
\gdef\@preamerr#1{%
  \begingroup
    \let\protect\relax
<*!autoload>
    \@latex@error{\ifcase #1 Illegal character\or
     Missing @-exp\or Missing p-arg\fi\space
     in array arg}\@ehd
</!autoload>
<autoload>  \@autoerr\@preamerr{#1}%
  \endgroup}
    \end{teX}
 \end{docCommand}

 \begin{docCommand}{@badlinearg}{}
 Occurs in |\line| and |\vector| command when a bad slope
 argument is encountered.
    \begin{teX}
\gdef\@badlinearg{%
<!autoload>  \@latex@error{%
<!autoload>       Bad \protect\line\space or \protect\vector 
<!autoload>       \space argument}\@ehb}
<autoload>  \@autoerr\@badlinearg}
    \end{teX}
 \end{docCommand}

 \begin{docCommand}{@parmoderr}{}
 Occurs in a float environment or a |\marginpar| when
 encountered in inner vertical mode.
    \begin{teX}
\gdef\@parmoderr{%
<!autoload>  \@latex@error{Not in outer par mode}\@ehb}
<autoload>  \@autoerr\@parmoderr}
    \end{teX}
 \end{docCommand}

 \begin{docCommand}{@fltovf}{}
 Occurs in float environment or |\marginpar| when there
 are no more free boxes for storing floats.
    \begin{teX}
\gdef\@fltovf{%
<!autoload>  \@latex@error{Too many unprocessed floats}\@ehb}
<autoload>  \@autoerr\@fltovf}
    \end{teX}
 \end{docCommand}

 \begin{docCommand}{@latexbug}{}
 Occurs in output routine.  This is bad news.
    \begin{teX}
\gdef\@latexbug{%
<!autoload>  \@latex@error{This may be a LaTeX bug}{Call for help}}
<autoload>  \@autoerr\@latexbug}
    \end{teX}
 \end{docCommand}

 \begin{docCommand}{@badcrerr}{}
    This error was removed and replaced by |\@nolnerr|.
 \changes{v1.0m}{1994/05/04}{Error message removed}
    \begin{teX}
\def\@badcrerr {\@latex@error{Bad use of \protect\\}\@ehc}
    \end{teX}
 \end{docCommand}

 \begin{docCommand}{@noitemerr}{}
 |\addvspace| or |\addpenalty| was called when not in
  vmode. Probably caused by a missing |\item|.
    \begin{teX}
\gdef\@noitemerr{%
<!autoload>  \@latex@error{Something's wrong--perhaps a missing %
<!autoload>      \protect\item}\@ehc}
<autoload>  \@autoerr\@noitemerr}
    \end{teX}
 \end{docCommand}

 \begin{docCommand}{@notprerr}{}
 A command that can be used only in the preamble
 appears after the command |\begin{document}|.
    \begin{teX}
\gdef\@notprerr{%
<!autoload>  \@latex@error{Can be used only in preamble}\@eha}
<autoload>  \@autoerr\@notprerr}
    \end{teX}
 \end{docCommand}

  \begin{docCommand}{@inmatherr}{}
 \changes{v1.0j}{1994/04/28}{Macro added}
 \changes{v1.1c}{1994/04/28}{Replaced \cs{noexpand} with \cs{protect}.}
    Issued by commands that don't work correctly within math (like
    |\item|). There is no real error recovery happening, e.g., the
    user might get additional errors afterwards.
    \begin{teX}
\gdef\@inmatherr#1{%
   \relax
   \ifmmode
<!autoload> \@latex@error{Command \protect#1 invalid in math mode}\@ehc
<autoload>  \@autoerr\@inmatherr#1%
   \fi}
    \end{teX}
  \end{docCommand}
%
% \begin{docCommand}@invalidchar}
% \changes{LaTeX2.09}{1993/09/19}
%     {(RmS) Error message for invalid input characters.}
% \changes{v1.0d}{1994/03/28}
%     {(DPC) Comment out (use catcode15 instead)}
%    An error for use with invalid characters.  This is commented
%    out, since we decided to use chatcode 15 instead.
%    \begin{teX}
%\def\@invalidchar{\@latex@error{Invalid character in input}\@ehc}
%    \end{teX}
% \end{docCommand}
 
 As well as the above error commands some error messages are directly
 coded to save space. The Messages alrerady present in \LaTeX2.09
 included: 

 |Environment --- undefined|\\
 Issued by |\begin| for undefined environment.

 |tab overflow|\\
  Occurs in |\= when| maximum number of tabs exceeded.

 |\< in mid line|\\
 Occurs in |\<| when it appears in middle of line.

 |Float(s) lost|\\
 In output routine, caused by a float environment or
 |\marginpar| occurring in inner vertical mode.


