\chapter{Introduction}

\section{Client Expectations}

During 2017, we have managed to meet most of the Client's expectations and where these were not met it was because of reasons beyond our control. During last year the Client kept tight control over milestones and in many cases setting new milestones with unrealistic targets. The Program has been divided into two portions, Program 1: Basements, Apartments and Mall and and Program 2, for the balalnce of the development including Hotel and Theme Experience Center.

A summary of where teh Site stands at the moment, will help better understand the interelationship between the Client's expectations and the year ahead.

\section{Basements}


\subsection{Mechanical} 

The Basements works are well advanced in the Mechanical Department with the status as follows:


We expect that all Mechancical works in the Basements will be completed by the end of March with the exception of cladding of piping. Cladding has been impeded by late deliveries of insulation, which was the single product that affected the works due to the blockade imposed onto Qatar by its neighbours.


\begin{longtable}{l c c c}
\toprule
Activity  & Total & Remaining & Planned Completion\\ 
          & Qty & Qty  & Date\\ 
          \midrule
Drainage  &     &    & 20 Jan, 2018 \\
H\&C      &     &    & 20  Jan, 2018\\
Chilled Water & &    & 20 Jan, 2018\\
FCU instalation & & & 20 Jan, 2018\\
Duct Installation & & & \\
Fire Rated Duct\footnote{Fire Rated duct installation, was disrupted due to the embargo, and Seagull experiencing financial issues and raw material deliveries. Mitigation steps included numerous meetings with Seagul management and the eventyual introduction of  a new Supplier (CVS).} & & &\\
Generator Flue & & &\\
\bottomrule
\end{longtable}

\paragraph{Constraints}  For lift lobbies, progress depends on Redco. At the moment these works are not carried out in parallel, but sequentially, so we expect delays in these activities.

\section{Electrical}

Electrical works, have fallen behind expectations, mainly because of the lack of Senior Engineers in this area. 


{\smaller
\begin{longtable}{l c c}
\toprule
Activity  & Total & Remaining \\ 
          & Qty & Qty\\ 
\midrule
Cabling &  &\\

Panels & &\\

Lights &  &\\

MEP works in rooms & &\\
\bottomrule
\end{longtable}
}

\subsection{Substations}

The Site has ten substations (one was additional works). The first nine substations are in an advanced state of construction. The following have been achieved.

\begin{longtable}{lll}
S/No            &No. Transformers &   \\
Substation-01   &  &\\
Substation-02   &  &\\
Substation-03   &  &\\
Substation-04   &  &\\
Substation-05   &  &\\
Substation-06   &  &\\
Substation-07   &  &\\
Substation-08   &  &\\
Substation-09   &  & \\
Substation-10   &  &\\
\end{longtable}


\paragraph{Outlook for Energizing the Substations}

Currently the Primary Substation (not in Conspel's scope), is almost completed. Kahraama are still to start testing with the current expectations is taht it will be completed by end of January, early Februray with energization expected to take place late March early April.

Our own substations we expect to submit to Kahraama for testing earlier than the PSS energization in order to gain time. This we expect to start as soon as the Power Factor Panels\footnote{Impacted by the blockade.} are delivered. The earliest will be end February 2018. With building energization end May 2018. 


\paragraph{Outlook for Power-on}

The outlook for Power-on for the Program 1, areas is less promising with my own expectation of Site readiness for end April, early May. Current constraints are the following:

\begin{description}
\item [Containment]  This has been going slower than planned mainly due to disruptions in the supply of cable trays and trunking, affected by the blockade. End of March 2018
\item [Cable Pulling] End of March 2018
\item [Terminations]  15 April 2018
\item [Panel Installation]  28 February 2018
\end{description}

Our bigger risk lies within this area and we are currently focusing on the above closely.








