\chapter{LaTeX Project}


\section{Requirements}

This Project ia about setting up a LaTeX installation on the AWS IaaS and the relevant other services that go with it. This Project is part of a larger project and will only focus on 

\section{Objectives}

\begin{itemize}
 \item  Setup LaTeX on the cloud and provide a UI to use it, similar to ShareLaTeX.
        
 \item Improve on user management to be able to create roles and authorizations.
 
 \item Handle all database interfaces and api.
\end{itemize}

\section{Workspace Management}

We expect the developer to be able to set up the Project on github.com and to be able to provide us with a full write up at the end of teh Project.

\section{Development Language}

Our preference is to use Node, Express, and related software for the development. The developer can provide details and motivaion for proposing to use other languages.


\section{Security}

The developer

\section{Modularization}

\section{Public/Private Projects}

The configuration module should extend to users, who want to make files and or pdfs public or private to download them or to get a tiny url that can email to their own group.

\section{Tests}

We expect the provider to develop a testing methodology and testing routines. We will provide two testing files in |.tex| format. Any help with the LaTeX part we can provide support. 

\section{References}

\href{https://github.com/aslushnikov/latex-online#compile-url}{aslushnikov}

\href{cloudlatex}{https://cloudlatex.io/}

\href{https://github.com/Khan/KaTeX}{katex}














