\chapter{Excel and SQLite3}

Many users will be using Excel Sheets to present and manipulate data. It is one of the most flexible
tools available for data manipulation.

\section{Exporting from Excel}

\section{Importing from Excel}

\section{Importing SQLite data into Excel}

The method escribed here worked for me using Windows 10, 64 bit.

\subsection{Steps}

\begin{enumerate}
\item Dowload the ODBC driver from \href{http://www.ch-werner.de/sqliteodbc/}{Christian Werner's} page. Select \texttt{sqliteodbc.exe}.

\item Install the driver by double-clicking the downloaded exe file. If you have any problem with permissions, just allow them.
\item  Start Excel: Select the \textit{Data} tab.
\item Select \textit{Read external data.}
\item Select \textit{From other sources}.
\item Select \textit{ODBC driver}
\item Follow through the wizard. 
\end{enumerate}

A nicer tutorial can be found at \href{https://menne-biomed.de/blog/sqlite-excel}{menne-biomed.de}

\section{Some useful tools}

Excel is a great tool, but slow and annoying sometimes. It is much quicker sometimes to just use command line tools for converting between excel files an csv. An excellent tool is \href{https://csvkit.readthedocs.io/en/1.0.3/index.html}{csvkit}. If you want to do more analysis \href{https://github.com/wireservice/agate}{agate} is a good python library.

Other command line tools is burntshushi's \href{https://github.com/BurntSushi/xsv}{xsv} and \href{https://github.com/dinedal/textql}{textql}. Even Microsoft has one Log Parser Studio.

Personally I prefer \texttt{csvkit}, as I can convert excel files faster, pipe the CSV into an editor for cleaning of the data, if need be and then importing into SQLite3. The conversion facility as well as the piping into other files is also very useful in 

If you need to import data from the shell, ensure you have curl, wget and git installed. All three can be very useful and can be used on windows as well. 